### 2017-11-20 README.Rmd
use_analysis()

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# refractoryANA

The goal of refractoryANA is to ...

### Citation

Please cite this compendium as:

> W. Francuzik, M. Worm, (`r format(Sys.Date(), "%Y")`). _Risk Factors and Managment of Refractory Anaphylaxis Research Compendium_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### Installation

You can install refractoryANA from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("wolass/refractoryANA")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2017-11-20 analysis/paper/paper.Rmd
use_analysis()

---
title: "Title Goes Here"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(refractoryANA)
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

# References

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

### 2017-11-21 analysis/paper/paper.Rmd
First part of the paper

@@
 ---
-title: "Title Goes Here"
+title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
 author:
-  - author 1
-  - author 2
+  - Wojciech Francuzik
+  - Margitta Worm, MD
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
     fig_caption: yes
@@
 abstract: |
-  Text of abstract
+  (maximum 200 words).
+
+  **Introduction:**  background; objectives; data sources; study eligibility criteria, participants, and interventions; study appraisal and synthesis methods; results; limitations; conclusions and implications of key findings.
+Authors are required to describe the significance of the topic under discussion.
+Areas covered - Authors are required to describe the research discussed and the literature search undertaken.
+
+  **Expert commentary:** - The author's expert view on the current status of the field under discussion. References must not be included in the abstract.
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock
 highlights: |
@@
 library(refractoryANA)
+require(magrittr,quietly = T)
+
 ```
+
+Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+
+**Corresponding author: **
+Prof. Dr. med. M. Worm margitta.worm@charite.de
+Phone: +49 30 450 529 005;	  Fax: +49 30 450 529 902
 
@@
 
-Here is a citation [@Marwick2017]
+Anaphylaxis is a life-threathening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties.
 
-# Background
+Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or repiratory compromise, or when the allergen is known or likely [@Sampson2006].
+Its pathogenesis  relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils.
 
-```{r demo, eval = FALSE}
-1 + 1
+The definition of _refractory_ anaphylaxis is not established.
+For the purpose of this review we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
+Unresponsiveness in this case is defined as a lack of expected normalization of clinical symptoms (i.e.
+rapid increase in blood pressure, bronchodilatation, tachycardia).
+
+It is important to distingush refractory anaphylaxis from recurrent anaphylaxis.
+In one case report authors used the term refractory anahpylaxis to describe 6 recurrent episodes of anaphylaxis assosciated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
+We did not include such reports in our analysis and advise on using the term "recurrent" anphylaxis istead of "refractory" in such cases.
+
+## Epidemiology
+
+The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
+Anaphylaxis in children is usally food dependent and less severe than episodes in adulthood.
+Cases of _refractory_ anaphylaxis are extremely rare.
+The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.
+
+## Risk factors for anaphylaxis
+
+Risk factors for anaphylaxis include: age, food elicitors (cow's milk and hen's egg) in children and drug induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013].
+Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigourous exercise (e.g.
+jogging) [@WormAllergy2017].
+Risk factors for _refractory_ anaphylaxis were not invesigated previously.
+
+## Management
+
+The managment algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.  
+Research in anaphylaxis treatment is difficult because it is an emergency condition. Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.
+
+We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current managment algorithms.
+
+<!-- We compared the successful procedures from the case reports to ??? -->
+
+<!-- ## Eligibility criteria -->
+<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale.
+-->
+
+<!-- ## Information sources -->
+<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->
+
+## Literature search strategy
+
+We have performed a database search in the "pubmed" database using the "RISmed"[@rismed] package for R statistical software [@R].
+The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017.
+The search was conducted on 16.11.2017 and returned 150 results.
+Published case reports of anaphylaxis treated with repeated doses of epinephrine  were included into the study (articles in English, Poltugese, French and Spanish were included into the analysis, one article in japanese was excluded).
+Additionally, case reports referenced in the included sources were added.
+We excluded reviews and papers not containg case description as well as cases which documented an appropriate response to injected adrenalin.
+Also cases describing Mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
+The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
+
+We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment.
+We used piloted polls to gather the data and all reports were screened once by the same researcher.
+
+All of the studies were case reports therfore the evidence level was classified as low.
+
+The review protocol was not established prior to the review and the review did not undergo preregistration.
+
+We have performed literature research identifying case reports of refractory anaphylaxis.
+All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included into this study.
+We investigated which interventions were reported as succesfull in the treatment of anaphylactic hypotension that was not reponding to injected (i.v.
+or i.m) adrenailn.
+Study outcomes were defined as time to elicit a response and mortality.
+Most of these reports described refractory anaphylaxis in a perioperative setting, but they also included cases of food and drug anaphylaxis.
+Suprisingly, no cases of insect venom anphylaxis were found among these reports.
+
+Outcomes: successful and unsuccesful managment of anaphylaxis was one of the most important outcomes we evaluated.
+This information was availible in all cases.
+Neurological deficits directly after the episode were also evaluated, but this measure might be biased.
+
+<!-- ## Data items -->
+<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->
+
+<!-- ## Risk of bias in individual studies -->
+<!-- Describe methods used for assessing risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis.
+-->
+
+<!-- ## Summary measures -->
+
+<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->
+
+<!-- ## Synthesis of results -->
+
+<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->
+
+<!-- ## Risk of bias across studies  -->
+
+<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->
+
+<!-- ## Additional analyses  -->
+
+<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->
+
+```{r demograph}
+source("../../R/data_handling.R")
+knitr::kable(demogr.tab,col.names = c("Elicitor","Age","Male fraction","Operative fraction"),digits = 1)
 ```
 
-# Methods
+# Risk factors of refractory anaphylaxis
 
-# Results
+The risk factors for refractory anaphylaxis are summarised in the table below.
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
+# Clinical symptoms of refractory anaphylaxis
+1. persistent hypotension - give methylene blue
+2. betablockers prevent adequate response to adrenaline - give glucagon
+
+# Therapy of refractory anaphylaxis
+
+Most of the treatment strategies were according to the guidelines of anaphylaxis management [source], but were supplemented by additional procedures that saved patients' lives.
+We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
+
+Ambulatory patients and not in hospital-setting were first treated with adrenaline wether patients experiencing anaphylaxis under anesthesia were usually first given intraveous volume replacement and 100% oxygen, followed by eithe epinephrine or alfa-adrenergic vasopressors or both.
+
+Epinephrine was given in 100% of cases in various doses.
+
+In Table X we provided therapeutic approaches used in the treatment of anaphylaxis.
+
+4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.
+
+```{r tabFatal}
+knitr::kable(fat)
 ```
 
-# Discussion
+# Methylene blue for the treatment of refractory hypotension
+
+`r which(d$What.helped_gr=="methylene blue") %>% length` reports (`r ((which(d$What.helped_gr=="methylene blue") %>% length)/length(d$What.helped_gr)*100) %>% round(1)`%) indicated a rapid increase in blood pressure after injection of methylene blue.
+It was both given as a bolus and as an infusion with the doses ranging from 0.5-2mg/kg.
+
+We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension with 100 mg bolus i.v.
+and follow it with an infusion of 50 µg/kg/min.
+As a vector a 100 mL of 5% dextrose can be used.
+
+We have no reports of methylene blue use in our registry
+
+It is important to remember that methylene blue injection may provoke false pulseoximetry readings as the blue dye mimics cyanosis.
+
+# Vasopressin as a supplementary vasopressor
+
+# Glucagon in anaphylaxis with concomitant beta blockers
 
 # Conclusion 
+<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article.
+Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->
 
-# Acknowledgements
+# Expert Commentary: 500-1000 words (included in overall word count).
+<!-- **To distinguish the articles published in the Expert Review series, authors must provide an additional section entitled ‘Expert Commentary’.
+This section affords authors the opportunity to provide their interpretation of the data presented in the article and discuss the developments that are likely to be important in the future, and the avenues of research likely to become exciting as further studies yield more detailed results.
+The intention is to go beyond a conclusion and should not simply summarise the paper.
+Authors should answer the following:** -->
+
+## What are the key weaknesses in clinical management so far?
+
+## What potential does further research hold? What is the ultimate goal in this field?
+
+## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?
+
+## Is there any particular area of the research you are finding of interest at present?
+<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->
+
+# Five-year view
+<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->
+
+# Key issues
+<!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->
+
+<!-- * Matrix-Assisted Laser Desorption/Ionization  Mass Spectrometry Imaging (MALDI-MSI) is able to provide a molecular insight of the samples, detecting the presence of a great variety of analytes directly on-tissue and showing their spatial distribution across the tissue section.
+-->
 
 # References
+<div id="refs"></div>
+<!-- **A maximum of 100 references is suggested.
+Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography.
+Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->
 
-### Colophon
+## Annotated bibliography
+<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+# Figures and Tables
+<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->
+
+```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
+source("../../R/prisma.R")
+knitr::include_graphics("../figures/prism_flow.png")
+```
+
+--==SOURCE SENTENCES==--
+asthma poses higher risk of anaphylaxis [@Koplin2011]
+less sun year-round is bound to higher food-induced anaphylaxis prevalence??? [@Koplin2011]
+low Vitamin D levels are might be associated with anaphylaxis? [@Koplin2011]
+
+The mainstay of therapy is early administration of intramuscular (IM) adrenalin with addi- tional therapies including intravenous (IV) fluids, steroids, and an- tihistamines after the administration of adrenalin [@Lazar2017].
+
+external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis??? [@Lazar2017]
+
+CASES:
+1.
+Male, 15yo, known milk allergy, food ANA, SPT +++, IgE+, Soy milk which previously contained cows milk.
+Epi 0.3 mg IM, repeated EPI + 0.6 mg, Salbutamol 5 mg intratracheal.
+500 ml iv NaCl.
+125 methylpred, Adrenalin 5mg/500 mL NACL - 10 µg/min.
+another Adreanlin 0.5 mg. ICU, Gastric lavage!!! 152 min after the exposure -> remission in 8 min. [@Lazar2017]
+
+The most important treatment of anaphylaxis includes early
+administration of IM adrenalin by an automatic injector1,2 [@Lazar2017]
+
+The patient's condition began to improve shortly after placement of a nasogastric tube and gastric drainage, suggesting that milkwas still being absorbed fromthe gastricmu- cosa and contributing to the anaphylaxis even after the correct therapywas given.
+[@Lazar2017]
+
+Similarly, although not previously reported, many allergists be- lieve that vomiting after food-induced IgE-mediated allergic reac- tion seems to end the allergic reaction. [@Lazar2017]
+
+Protocols for therapy of anaphylaxis have been sug-
+gested1,2,4,5; however, none of these protocols include gastric drainage through a nasogastric tube as ameasure to decrease aller- gic absorption from the gastrointestinal tract in cases of severe anaphylaxis due to food.
+[@Lazar2017]
+
+Our case also shows the importance of early ventilation with
+supplemental 100%oxygen in severe, near-fatal anaphylaxis. [@Lazar2017]
+
+The most common causes of anaphylaxis are food and medications; however, Hymenoptera stings, radiocontrast media (RCM), and iatrogenic procedures SIT, [@Coop2017]
+
+Patients taking anti- hypertensive medications, specifically angiotensin converting enzyme (ACE) inhibitors or beta-blockers, represent a popula- tion in whom a possibly increased risk for anaphylaxis has been debated.3 [@Coop2017]
+
+Although the lifetime prevalence of anaphylaxis in the general population is estimated to be approximately 7% and possibly up to 10% to 12%, [@Coop2017]
+
+the prevalence of anaphylaxis in patients taking ACE inhibitors and beta-blockers is unknown.4 [@Coop2017]
+
+Anaphylaxis is classically initiated by allergen cross-linking of
+IgE on mast cells and basophils, leading to degranulation and release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes.5,6 [@Coop2017]
+
+Release of these mediators activates inflammatory pathways, including the complement and kallikrein-kinin systems and the clotting cascade.
+Histamine release, in particular, causes vasodilation leading to increased vascular permeability and decreased peripheral vascular resis- tance.2 [@Coop2017]
+
+Decreased peripheral vascular resistance leads to activa- tion of the renin-angiotensin system, a compensatory mechanism blocked by ACE inhibitors, thus theoretically leading to inten- sified anaphylaxis.
+[@Coop2017]
+
+Histamine release also increases cardiac rate, cardiac contractility, and bronchoconstriction.2,5 Therefore, it is suggested that beta-blockers mask cardiac signs of anaphylaxis and lead to unopposed alpha-adrenergic activity, causing severe bronchoconstriction.
+[@Coop2017]
+
+Furthermore, management of anaphylaxis with concomitant beta-blocker use is complicated by the possi- bility of a blunted response to epinephrine and/or inhaled bronchodilators.7 [@Coop2017]
+
+Also animal studies suggest that ACE in- hibitors and beta-blockers together may enhance mast cell priming.8 [@Coop2017]
+
+Nitric oxide also plays a role in anaphylaxis. Nitric oxide can
+inhibit the release and effects of catecholamines.
+It causes vaso- dilation by increasing the activation of guanylyl cyclase that can result in smooth muscle relaxation via cyclic guanosine mono- phosphate.9 [@Coop2017]
+
+Methylene blue is an inhibitor of guanylyl cyclase and has been used as a treatment for refractory anaphylaxis.10 [@Coop2017]
+
+There is evidence that glucagon is also helpful in refractory anaphylaxis especially for patients on beta-blockers.
+Glucagon activates adenylyl cyclase directly and can bypass the beta blockade.11,12 [@Coop2017]
+
+Additionally, higher doses of epinephrine may be needed to overcome the beta blockade for refractory anaphylaxis; however, beta-blockers prevent vasodilation, leaving unopposed alpha vasoconstriction that may result in a hypertensive reaction.12,13 [@Coop2017]
+
+Jacobs et al14 reported on 2 patients who developed anaphylaxis while taking propranolol, a beta-blocker.
+Both of these patients developed hypotension and were refractory to epinephrine.
+The authors stated that propranolol can be impli- cated as a potentiating factor in anaphylaxis because of its antagonistic action on the beta-adrenergic sympathetic system.
+[@Coop2017]
+
+A publication in 1983 documented 5 patients who had re-
+fractory anaphylaxis while taking propranolol.15 Another case described a case of fatal bronchospasm and anaphylaxis in a patient using beta-blocker (timolol) eye drops.16 Finally, Hayhoe et al17 described a 42-year-old woman who had severe anaphy- laxis from labetalol use.
+She had persistent bradycardia, elevated tryptase levels, and hypotension that was refractory to epineph- rine and norepinephrine.
+Eventually she responded to enox- imone that is a nonadrenergic inotrope.
+[@Coop2017]
+
+More recently, Aggarwal et al27 reported that beta-blocker premedication did not increase the frequency of allergic re- actions to coronary computed tomography angiography even among patients who had a past history of RCM reactions.
+The incidence of an allergic reaction in patients who received beta- blockers was 45 of 23,867 (0.19%) compared with those pa- tients who did not receive beta-blockers, which was 9 of 5232 (0.17%).
+There was no statistical difference between the 2 groups.
+[@Coop2017]
+
+Overall, there is not enough data in the medical literature regarding venom allergy and anaphylaxis. [Coop2017]
+
+A case report in 1981 docu- mented anaphylaxis refractory to epinephrine in a patient taking propranolol and receiving a maintenance immunotherapy in- jection to grass pollen.34 [@Coop2017]
+
+There have been some reports of death of patients receiving immunotherapy while taking beta-blockers or ACE inhibitors.
+In a study published in 1987, Lockey et al35 reported 2 fatalities of immunotherapy patients who were taking beta-blockers.
+Reid et al36 published a study with 1 death of a patient taking a beta-blocker while receiving immunotherapy in a survey of allergists from 1985 to 1989.
+There were 17 re- ported immunotherapy fatalities during this survey period.
+[@Coop2017]
+
+Another study over a 12-year period collected data on fatal
+reactions to immunotherapy and skin testing.37 There were 41 immunotherapy fatal reactions and only 1 patient was found to be taking an ACE inhibitor.
+None of the patients were taking beta-blockers.
+Lockey et al38 again published a review article in 2001 reporting that patients on beta-blockers were at risk for systemic reactions and fatalities from allergen immunotherapy.
+[@Coop2017]
+
+A study in 2006 by Amin et al39 described a near fatal reaction
+in a patient taking a beta-blocker and receiving allergen immu- notherapy.
+Interestingly, there were also 2 patients with near fatal reactions who were taking angiotensin receptor blockers while receiving immunotherapy.
+[@Coop2017]
+
+Rank et al45 evaluated 29 immunotherapy systemic reactions
+in a study of 338 immunotherapy patients over a 2-year period.
+ACE inhibitors and beta-blockers did not increase the odds of having a systemic reaction to subcutaneous allergen immuno- therapy.
+Additional studies of Hymenoptera venom allergic pa- tients while taking beta-blockers and ACE inhibitors were conducted.
+These studies found that the patients taking cardio- vascular medications did not have an increased risk of systemic reactions to build-up venom immunotherapy.
+[@Coop2017]
+
+Finally, Müller and Haeberli49 undertook a prospective study
+of patients with cardiovascular disease who were taking beta- blockers and receiving venom immunotherapy.49 Of the 25 patients on beta-blockers during immunotherapy, 3 (12%) developed systemic reactions, compared with 23 (16.7%) of 117 patients not taking beta-blockers.
+No severe reactions were re- ported in the patients taking beta-blockers and receiving venom immunotherapy.
+The authors concluded that beta-blockers may be indicated in patients with cardiovascular disease who are receiving venom immunotherapy.
+[@Coop2017]
+
+However, for those patients with cardiovascular disease, beta-blockers and ACE inhibitors have been shown to decrease mortality and increase life expectancy.
+Consideration should be given for patients with cardiovascular disease and venom immunotherapy to remain on their beta-blockers and ACE in- hibitors.49 We recommend that these medications be continued for patients who need them on venom immunotherapy.[@Coop2017]
+
+The authors state that “current recommendations are to start with an infusion of epinephrine for unresponsive anaphylaxis; and if there is re- fractory hypotension, then add dopamine as a second vaso- pressor.”1 The literature supports the use of epinephrine infusion in refractory anaphylactic shock because it reverses most of the cardiovascular effects of anaphylaxis.3 At low doses, dopamine exerts its effects on the b-adrenergic receptors, causing an in- crease in myocardial function, heart rate, and some vasodilata- tion.
+However, there is no clear evidence for the use of dopamine as a second-line vasopressor in anaphylactic shock.
+This is particularly important given that dopamine has been associated with a higher risk of mortality and major adverse events compared with epinephrine and norepinephrine.4,5 The literature suggests that norepinephrine or vasopressin may be appropriate second-line agents.
+Norepinephrine acts primarily on the a-adrenergic receptors, whereas vasopressin acts on the V1 re- ceptors, both leading to vasoconstriction.2,6 Methylene blue has also been used with good success in catecholamine-refractory vasoplegic anaphylactic shock.7 [@Alquarashi2016]
+
+This complication is of particular concern in young children given their anatomy and higher susceptibility to significant obstruction with even a mild degree of airway edema.
+Because of its rapid effectiveness, ease of administration, and reasonable safety profile, inhaled nebulized epinephrine should be considered in children with anaphylaxis who present with stridor or other signs of upper airway obstruc- tion.9,10 Health care professionals should consider this intervention before proceeding to advanced airway measures, especially in prehospital settings.
+It should be noted that inhaled nebulized epinephrine acts primarily in the upper airway with minimal sys- temic effects.
+Therefore, it is not an alternative but rather adjunc- tive treatment of anaphylaxis with injectable intramuscular epinephrine.9 [@Alquarashi2016]
+
+Finally, we would like to suggest a correction in the published glucagon dose.
+In children, the loading dose is 20 to 30 µg/kg (maximum, 1 mg), and the infusion dose is 5 to 15 µg/min rather than the dose recommended in the 2015 update [@Alquarashi2016]
+
+A 36 year old female developed anaphylaxis intraopera- tively while undergoing a bilateral knee arthroscopy under general anaesthesia.The diagnosis of anaphylaxis was made and treatment initiated.
+Intravenous adrenaline was administered in 0.5 mg doses every two minutes.
+She also received 6 mg of etilefrine 1 mg of atropine and 2 litres of Ringer’s lactate.
+Despite 2 mg of adrenaline and 1 mg atropine intravenously, she remained hypotensive with a bradycardia.
+She was also given 200 mg of hydrocortisone, 25 mg of promethazine and 50 mg of ranitidine intravenously.
+As a last resort, 1 IU of ornipressin was administered intravenously.
+This resulted in restoration of arterial blood pressure and an increase in heart rate.
+Subsequent doses of 50-100 μg of adrenaline resulted in a significant increase in blood pressure and heart rate.
+Forty five minutes later the reaction reoccurred and once again
+responded poorly to treatment with boluses of
+adrenaline.
+Administration of 1 IU of ornipressin restored haemodynamics promptly once again.
+An adrenaline infusion of 0.02 to 0.05 μg/kg/min and another 5 litres of intravenous fluid was required over the next six hours after which she completely stabilised.
+She was discharged from hospital three days later.
+mmunoCAP testing was performed six weeks after discharge from hospital and was positive for latex specific IgE with a value of 5.06 IU/ml.
+ImmunoCAP and CAST tests are not available for the drugs that were administered to her and in view of the severity of the reaction, it was not considered safe to perform skin prick testing or a drug challenge.
+While the anaphylaxis may possibly have been due to latex, the cause remains unknown.
+ Vasopressin or ornipressin should be considered for the treatment
+of adrenaline-resistant anaphylactic shock before resuscitation is discontinued. [@Seedat2014]
+
+USE OF METHYLENE BLUE IN CHEMOTHERAPY-INDUCED REFRACTORY ANAPHYLACTIC SHOCK 
+Learning Objectives: Distributive shock is common in the critically ill patient.
+Resuscitation with fluids and catecholamine agents have become standard of care.
+Methylene blue (MB) has been proposed as an alternative treatment of shock.
+Methods: A 67-yr- old male with history of chronic lymphocytic leukemia on cycle 2 of bendamustine and rituximab, with recent admission for rash and hypotension attributed to allopurinol after cycle 1, developed shortness of breath, chest tightness, and abdominal discomfort while receiving chemotherapy.
+Treatment was stopped and the patient received fluids, steroids, and diphenhydramine.
+He became tachycardic, hypertensive, and hypoxic.
+Te patient was intubated for airway protection.
+Physical exam revealed a diffuse, blanching erythematous rash.
+Upon arrival to the intensive care unit, he remained normotensive and extremities were well-perfused.
+He received additional fluids, steroids, and antihis- tamines.
+Several hours later, he developed hypotension necessitating initiation of vaso- pressors.
+Elevated tryptase level was consistent with clinical diagnosis of anaphylactic shock.
+Shock was refractory to maximal support on four vasopressors.
+A four-hour MB infusion was started at 0.5 mg/kg/hr and titrated to 2 mg/kg/hr.
+Shock improved and the patient was weaned down to three vasopressors.
+After a short period of time, he again became hypotensive on maximal vasopressor support.
+Another MB infusion was started and the patient was weaned down to two vasopressors.
+He did not require additional doses of methylene blue.
+Te patient improved and was discharged home.
+Results: Reported use of MB for anaphylaxis is limited.
+Effect of MB on vasodilation has mostly been studied in sepsis.
+Increased nitric oxide (NO) production is associated with loss of vascular tone in vasodilatory shock.
+MB inhibits NO-induced guanylate cyclase activa- tion and increases smooth muscle tone.
+MB may be an effective alternative treatment to restoring vascular tone in anaphylactic shock as evident in our case.
+Studies are needed to determine ideal dosage, duration of treatment, and effect on mortality.[@Tsai2016]
+
+The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]
+
+The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]
+
+ The exact prevalence of anaphylaxis in the general population is
+not known, but it is widely believed to be an underrecognized and undertreated medical emergency.1e9 [@Fineman2015]
+
+? Epinephrine in appropriate doses is safe, and there are no abso- lute contraindications for its use in treating anaphylaxis.
+? Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.
+? Epinephrine administration is not only indicated for use in anaphylaxis but also for a severe allergic reaction or for patients
+identified as being at risk of anaphylaxis.
+? It is not necessary for the NIAID/FAAN criteria to be met to administer epinephrine.
+? Antihistamines and glucocorticoids are not indicated as first-line treatment for anaphylaxis but may be administered after the
+administration of epinephrine if considered to be appropriate in the judgment of the treating professional.
+? Patients treated in the emergency setting for anaphylaxis or for severe allergic reactions or those who are at risk of a future
+event should be provided with a prescription for epinephrine autoinjectors and an action plan for their use before discharge.
+? Patients should be referred to an allergist to assist with diagnosis confirmation, trigger identification, and continued outpatient
+management.[@Fineman2015]
+
+MMAS with recurrent unprovoked life-threatening anaphylaxis responding to treatment with omalizumab.h 6 multisystem reactions
+over the course of the previous 8 months.
+Her typical symptoms peaked within 5 to 10 minutes of onset and included cutaneous (urticaria, swelling), respiratory (cough, shortness of breath, chest tightness, dysphagia, dysphonia), gastrointestinal (vomiting, cramping abdominal pain), and cardiovascular involvement.
+Blood pressure was recorded at 80/40 mmHg by paramedics during 1 reaction and loss of conscious occurred on 1 occasion.
+All 6 episodes had been treated in the emergency department and required 2 to 3 doses of epinephrine.
+The patient was not taking angiotensin- converting enzyme inhibitor, b blocker, nonsteroidal anti- inflammatory, or opioid medication.
+Typical triggers, including foods, drugs, latex, insect stings, exercise, menstruation, and alcohol, were ruled out and skin testing was noncontributory.
+[@Jagdis2014]
+
+5 Patients treatet with omalizumab  with refractory anaphylaxis with SM / MCAS. Omalizumab is a humanized murine monoclonal antibody that
+complexes with free IgE in serum, preventing IgE from binding to FcεR1 on mast cells and basophils and resulting in downregulation of FcεR1.
+There are 4 case reports of 5 patients on the use of oma- lizumab in the spectrum of mast cell activation disorders (Table 1).5e8 [@Jagdis2014]
+
+This is the third reported case of omalizumab use in MMAS and suggests thatomalizumabmay be an effective adjunct to therapy in patients with MMAS and life-threatening reactions re- fractory to maximal medical therapy.[9] Molderings GJ, Raithel M, Kratz F, et al.
+Omalizumab treatment of systemic mast cell activation disease: experiences from four cases.
+Intern Med.
+2011; 50:611e615.
+[@JAgdis2014]
+
+The AAAAI/ACAAI Guidelines show info about refractory anaphylaxis [@Simons2014]
+The three guidelines on the management of refractory anaphylaxis: [@Simons2014] 
+Management of refractory anaphylaxis
+WAO: intubation; ventilation; IV vasopressors; glucagon; anticholinergic; transfer to hospital (preferably to an emergency medicine, critical care medicine, or anesthesiology) team for ventilatory and inotropic support; checklist of needed items; Table 6
+AAAAI: vasopressors; dopamine; give vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension; transfer to hospital; glucagon; atropine; methylene blue; includes checklist of supplies and equipment; Figures E2, E3
+EAACI: glucagon
+
+Management of anaphylaxis refractory to initial treatment The guidelines differ in emphasis on refractory anaphylaxis treatment. The WAO Guidelines stress the importance of
+prompt initial treatment to prevent escalation of symptoms [2].
+They suggest that if possible, patients with anaphylaxis refractory to epinephrine, supplemental oxygen, IV fluids, and second-line medications should be transferred to the care of a specialist team for ventilatory and inotropic support and continuous electronic monitoring [2].
+The AAAAI/ACAAI Guidelines provide details about interven- tions for cardiopulmonary arrest, airway management, and IV administration of vasopressors including epinephrine, dopamine, and vasopressin [3].
+The EAACI Guidelines in- clude brief specific instructions about when to call for In- tensiveCareUnit support [4](Table4).
+Studies relevant to refractory anaphylaxis treatment
+are of interest.
+A Cochrane review of randomized con- trolled trials (RCT) in more than 20,000 critically ill pa- tients with distributive shock supports administration of crystalloids such as 0.9% saline, because administration of colloids such as albumin or hetastarch did not correlate with increased survival [82].
+Methylene blue administration for vasoplegia in anaphylaxis refractory to epinephrine and IV fluid resuscitation is based on case reports and extrapolation from use in other forms of shock [83].
+[@Simons2014]
+
+Management of anaphylaxis refractory to initial treatment Even in high-resource countries, optimal treatment of re- fractory anaphylaxis is not available universally; for ex- ample, in remote, inaccessible, or impoverished areas or in specific situations such as anaphylaxis on airplanes.
+In limited-resource situations, lack of availability of basic es- sentials such as epinephrine, supplemental oxygen and IV fluid resuscitation is more critical than lack of second-line medications such as antihistamines and glucocorticoids.
+Lack of availability of advanced life-support management can be a major barrier to survival [72-76,128] (Table 8).
+In any limited-resource situation, resuscitation efforts pro- longed over hours using a hand-held bag valve mask (manual resuscitator) are often successful in anaphylaxis [2] (Table 8).
+In mid- and low-resource countries, striving to ensure
+more consistent availability of medications, supplies, and equipment for anaphylaxis treatment is an important goal [2,5-7].
+The World Health Organization has devel- oped a tool kit containing evidence-based guidelines and  a framework for quality improvement in the hospital care of critically ill children in such environments [126], where despite many obstacles, improvements can be documented [130].
+[@Simons2014]
+
+DIAGNOSTIC CRITERIA FOR ANAPHYLAXIS
+Anaphylaxis remains a clinical diagnosis based on probability and pattern recognition. It has no universally accepted clinical definition [1,5]. [@Kemp2014]
+
+In clinical practice, however, waiting until the
+development of multiorgan symptoms is risky because the ultimate severity of an anaphylactic reaction is difficult to predict at its onset.
+Anaphylaxis occurs as part of a clinical continuum that can begin with relatively minor symptoms such as itchy skin, eyes, or nose and rapidly progress to a life- threatening respiratory or cardiovascular reaction, or both.
+Urticaria and angioedema are the most common manifestations [7–9].
+Cutaneous findings may be delayed or absent in rapidly progressive anaphylaxis.
+The next most common manifestations of anaphylaxis are respiratory symptoms, followed by dizziness, syncope, and gastrointestinal symptoms.
+The more rapidly anaphylaxis occurs after exposure to an offending stimulus, the more likely the reaction will be severe and potentially life- threatening [10,11 [@Kemp2014]
+
+No high-quality studies at present permit recommendations concerning the use of
+intravenous epinephrine in anaphylaxis [18] Parameters and international 
+guidelines support its use in refractory anaphylaxis [20–23] (Table 1). BUt it can cause arrythmias that are lethal. [@Kemp2014]
+
+MANAGEMENT OPTIONS AFTER INTRAMUSCULAR EPINEPHRINE
+Options after epinephrine injections include supple- mental oxygen and ß2-agonists, volume replace- ment, and intravenous epinephrine.
+Supplemental
+
+We present the case of a 43-year-old woman with severe ana- phylaxis unresponsive to epinephrine.
+Physical examina- tion revealed marked respiratory distress, raised oral lesions, and altered mental status but lacked hypotension.
+After infusion of methylene blue, symptom resolution occurred almost immediately, and intubation was spared.
+Side effects were minimal.
+We propose methylene blue as a safe treat- ment option for refractory anaphylaxis, whether with or without hypotension.[@Bauer2013]
+
+Anaphylaxis can occur after an immunoglobin E–
+mediated reaction to food, drugs, insects, or other stimuli.
+It is termed idiopathic when the cause is unknown.
+The mainstay of treatment is epinephrine, followed by oxygen, fluids, antihistamines, inhaled β-agonists, and glucocortico- steroids.
+In cases of anaphylaxis “resistant” to epinephrine, alternatives must be used.
+[@Bauer2013]
+
+Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
+10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - return of spontaneaous circulation.
+[@Wang2016] This case shows how important it is to get rid of the alergen in refractory anaphylaxis.
+Traetment with ECMO.
+
+Abstract Anaphylaxis is a life-threatening reaction treated primar-
+ily with epinephrine.
+Methylene blue, a competitive inhibitor of guanylate cyclase, interferes with the vasodilatory actions of nitric oxide.
+It has recently been proposed by the Joint Taskforce on Practice Parameters as an alternative treatment for anaphylaxis with hypotension that is not responsive to classical therapy.[@Bauer2013]
+
+Because anaphylaxis is potentially fatal, treatment alternatives for cases refractory to epinephrine are essential.
+Methylene blue has been shown to be a life-saving option with few adverse effects [3,8].
+It should be considered a safe and effective option for patients, with or without hypoten- sion, whose anaphylactic reactions are not responsive to epinephrine (Fig.
+2).
+Cindy[@Bauer2013]
+
+Methylene blue has been successfully used in cases of
+refractory shock due to vasoplegia after extracorporeal cir- culation, sepsis, and anaphylactic shock3,6,7,14
+; however, its mechanism of action that results in shock reversal has yet to be fully elucidated. Excessive activation of the NO-cGMP pathway occurs in refractory shock,6,14
+and studies indi- cate that methylene blue inhibits guanylate cyclase, which interrupts this pathway and reverses vasodilation, caus- ing hemodynamic improvement.7,11 [@Silva2017]
+
+Methylene blue is a selective inhibitor of the NO-cGMP pathway7 and does
+not have the side effects of nonselective inhibitors of NO release.3,7
+The dose used in cases of shock is 1–2 mg/kg,3,7
+but additional doses or continuous infusion may be required due to its short half-life.7,14
+Regardless, there is no solid evi-
+dence to support the use of methylene blue as a single drug in cases of anaphylaxis; it is rather applied as an adjuvant in treatment with catecholamines.3,6,15
+The usual treatment was
+used in this case with an unsatisfactory effect; the patient remained in shock for approximately 15 minutes even with high doses of catecholamines.
+Methylene blue administra- tion was associated with an immediate improvement in hemodynamics and tissue perfusion.
+The immediacy of improvement suggests that the effect was due to its use and not a late and unique effect of the catecholamines [@Silva2017]
+
+While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000---20,000 anesthesia pro- cedures, it is responsible for 3---10% of the perioperative fatalities. [@Nag2017]
+
+Allergic reactions to anesthetic drugs usually occur within 10 min of the drug exposure but can also occur as late as 30 min to several hours later.4 However, more than 90% of reactions evoked by intravenous drugs occur within 3 min of its administration.5 In[@Nag2017]
+
+Bradycardia is an uncommon presentation during anaphylaxis. In our patient, premedication with beta blocker and severe hypoxia could be the probable reason for this uncommon presentation.[@Nag2017]
+
+# Algorithm for the treatment of refractory anaphylaxis.
+[@Bauer2013]
+make it better
+
+# Firs sings in anesthesia setting of anaphylaxis
+1. In our two cases ETco2 collapse was the first sign preceding cardiac arrest. Several authors have underlined
+the value of this early clinical sign for the
+suspicion of anaphylactic shock, even in the absence of
+any cutaneous manifestations. [@Baumann2009]
+
+Epinephrine-resistant anaphylactic shock is a
+strong incentive to search for possible alternative
+treatments.8 Methylene blue,9 glucagon,10 and
+-agonists11 have recently been proposed as alternative
+therapeutic options. Successful results have
+also been reported using arginine-vasopressin12
+or vasopressin analogues (terlipressin)13 in epinephrineresistant
+anaphylactic shock resuscitation. Argininevasopressin
+is not available in France, therefore we
+used terlipressin. Cardiopulmonary bypass and extracorporeal
+support have also been successfully used in
+one case after 60 min of futile resuscitation.14 However,
+prompt resuscitation, use of terlipressin, and
+circulatory support were ineffective in our cases. [@Baumann2009]
+
+<!-- ### Colophon -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
 
@@
 # which R packages and versions?
-devtools::session_info()
+# devtools::session_info()
 ```
@@
 # if your Rmd is not in analysis/paper/
-git2r::repository("../..")
+# git2r::repository("../..")
 ```

### 2017-11-21 R/analysis.R

@@ -1,0 +1,37 @@
+# The search was performed on 16.11.2017
+require(RISmed)
+require(magrittr)
+res <- EUtilsSummary("refractory anaphylaxis",
+type="esearch",
+db="pubmed",
+datetype='pdat',
+mindate=1950,
+maxdate="2017/11/16",
+retmax=500)
+QueryCount(res)
+t<-ArticleTitle(EUtilsGet(res))
+tdf <-data.frame(No=1:length(t),t)
+first <- tdf[c(1,3,9,14,26,30,36,39,43,44,49,52,54,67,68,69,71,72,73,74,75,76,87,88,89,
+    90,91,92,93,97,98,100,106,108,114,120,122,123,124,125,131,132,134,141,149),]
+second <- tdf[c(3,9,14,26,30,44,54,67,69,71,72,73,74,75,76,87,
+              90,91,92,93,97,108,114,132,134),]
+fulltext_missing <- c(68,120,122,123,124,125) %>% length
+japanese <- 88 %>% length()
+no_adrenaline <-98%>% length()
+not_refractory <- c(106, 97)%>% length()
+guineapigs<- c(141,149) %>% length()
+no_case_rep <- 131 %>% length()
+
+# Include information from the anaphylaxis registry
+#load("~/Documents/anaphylaxis/data3.R")
+## See multiple adrenalin cases
+#require(magrittr)
+#data3$d_560_adren2_v5 %>% summary
+#data3[data3$d_560_adren2_v5=="yes",15:51] %>% summary
+#data3$q_142_fatal_treatment_v5%>% summary
+#data3[data3$q_142_fatal_treatment_v5=="yes",51:100] %>% summary
+
+
+
+
+

### 2017-11-21 R/data_handling.R

@@ -1,0 +1,152 @@
+######## Load the data ########
+d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
+
+######## prepearing the data frame ######
+mean(d$age,na.rm = T)
+d$Elicitor_gr <- rep(NA,length(d$Elicitor))
+d$Elicitor_gr[c(grepl("cont",d$Elicitor),24)] <- "contrast"
+d$Elicitor_gr[c(4,7,13,19,20,21,38,39)] <- "unknown"
+d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,23,27,28,29,30,31,32,33,34,35,36,40,41)] <- "drug"
+d$Elicitor_gr[c(1,3,5,15,25,37,42)] <- "other"
+
+d$Time.after.exposure
+d$Time <- c(5,5,10,NA,5,NA,5,NA,NA,NA,NA,NA,NA,NA,15,NA,NA,5,5,NA,NA,5,5,0.2,5,5,0.5,NA,5,20,5,5,10,5,5,5,15,NA,NA,0.5,0.5,5,NA)
+
+d$Situation_gr <- c("OP",
+                    "OP",
+                    "OP",
+                    "ambulatory",
+                    "not hospital",
+                    "chemo",
+                    "OP",
+                    "unknown",
+                    "unknown",
+                    "unknown",
+                    "OP",
+                    "OP",
+                    "OP","unknown",
+                    "not hospital",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "unknown",
+                    "unknown",
+                    "CT",
+                    "chemo",
+                    "OP",
+                    "ambulatory",
+                    "OP",
+                    "OP",
+                    "ambulatory",
+                    "not hospital",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP",
+                    "OP")
+
+# One case needs removal due to non-refractory character: LAMB
+d <- d[-4,]
+d <- d[-13,]
+d <- d[-27,]
+d <- d[-34,]
+
+
+########### # The demographics table  ######
+
+demogr.tab <- data.frame(Age=d$age %>% split(d$Elicitor_gr) %>%
+               lapply(mean,na.rm=T) %>%
+               unlist,
[…]

### 2017-11-21 R/prisma.R

@@ -1,0 +1,49 @@
+# Prisma flow diagdramm
+require(DiagrammeR)
+
+create_graph() %>%
+  add_node(x = -1.2, y = 4,
+           label="Records identified through\ndatabase searching\n(n = 150)",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 1.2, y = 4, label="Additional records idientified\nthrough references\n(n = 18)",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = 0, y = 3, label="Records after duplicates removed\n(n = 161)",
+           shape="rectangle",
+           width=2.5) %>%
+  add_node(x = 0, y = 2, label="Records screened\n(n = 161)",
+           shape="rectangle",
+           width=2) %>%
+  add_node(x = 3, y = 2, label="Records excluded\n(n = 115)",
+           shape="rectangle",
+           width=2) %>%
+  add_node(x = 0, y = 1, label="Full-text articles assessed\nfor eligibility\n(n = 46)",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 3\nnot a case report = 3\nnot anaphylaxis = 2\nanimal study = 2\njapanese language = 1",
+           shape="rectangle",
+           width=2,
+           height=1) %>%
+  add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 37)",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_edge(1,3) %>%
+  add_edge(2,3) %>%
+  add_edge(3,4) %>%
+  add_edge(4,6) %>%
+  add_edge(4,5) %>%
+  add_edge(6,7) %>%
+  add_edge(6,8) %>%
+  select_edges() %>%
+  add_global_graph_attrs(
+    attr = "splines",
+    value = "spline",
+    attr_type = "graph") %>%
+  # render_graph() %>%
+  export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/prism_flow.png", file_type = "png", title = NULL,
+               width = NULL, height = NULL)

### 2017-11-23 analysis/paper/paper.Rmd
Changes to the main file, citations,graph

--- 2017-11-20 analysis/paper/paper.Rmd
@@
 bibliography: references.bib
-csl: journal-of-archaeological-science.csl
+csl: mee.csl
 abstract: |
-  Text of abstract
+  (maximum 200 words).
+
+  **Introduction:**  background; objectives; data sources; study eligibility criteria, participants, and interventions; study appraisal and synthesis methods; results; limitations; conclusions and implications of key findings.
+Authors are required to describe the significance of the topic under discussion.
+Areas covered - Authors are required to describe the research discussed and the literature search undertaken.
+
+  **Expert commentary:** - The author's expert view on the current status of the field under discussion. References must not be included in the abstract.
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock
 highlights: |
@@
 
-Here is a citation [@Marwick2017]
+Anaphylaxis is a life-threathening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties.
 
-# Background
+Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or repiratory compromise, or when the allergen is known or likely [@Sampson2006].
+Its pathogenesis  relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils.
+*Anaphylaxis is classically initiated by allergen cross-linking of IgE on mast cells and basophils, leading to degranulation and release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes.5,6 [@Coop2017]*
 
-```{r demo, eval = FALSE}
-1 + 1
+*Anaphylaxis remains a clinical diagnosis based on probability and pattern recognition. It has no universally accepted clinical definition [1,5]. [@Kemp2014]*
+
+The definition of _refractory_ anaphylaxis is not established.
+For the purpose of this review we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
+Unresponsiveness in this case is defined as a lack of expected normalization of clinical symptoms (i.e.
+rapid increase in blood pressure, bronchodilatation, tachycardia).
+
+It is important to distingush refractory anaphylaxis from recurrent anaphylaxis.
+In one case report authors used the term refractory anahpylaxis to describe 6 recurrent episodes of anaphylaxis assosciated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
+We did not include such reports in our analysis and advise on using the term "recurrent" anphylaxis istead of "refractory" in such cases.
+
+## Epidemiology
+
+The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
+Anaphylaxis in children is usally food dependent and less severe than episodes in adulthood.
+Cases of _refractory_ anaphylaxis are extremely rare.
+The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.
+
+*The exact prevalence of anaphylaxis in the general population is not known, but it is widely believed to be an underrecognized and undertreated medical emergency.1e9 [@Fineman2015]*
+
+## Risk factors for anaphylaxis
+
+Risk factors for anaphylaxis include: age, food elicitors (cow's milk and hen's egg) in children and drug induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013].
+Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigourous exercise (e.g.
+jogging) [@WormAllergy2017].
+Risk factors for _refractory_ anaphylaxis were not invesigated previously.
+
+## Management
+
+The managment algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.
+The mainstay of therapy is early administration of intramuscular (IM) adrenalin [@Lazar2017].
+ However, research in anaphylaxis treatment is difficult because it is an emergency condition and placebo controlled studies are impossible.
+Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.
+
+We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current managment algorithms.
+
+<!-- We compared the successful procedures from the case reports to ??? -->
+
+<!-- ## Eligibility criteria -->
+<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale.
+-->
+
+<!-- ## Information sources -->
+<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->
+
+## Literature search strategy
+
+We have performed a database search in the "pubmed" database using the "RISmed"[@rismed] package for R statistical software [@R].
+The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017.
+The search was conducted on 16.11.2017 and returned 150 results.
+Published case reports of anaphylaxis treated with repeated doses of epinephrine  were included into the study (articles in English, Poltugese, French and Spanish were included into the analysis, one article in japanese was excluded).
+Additionally, case reports referenced in the included sources were added.
+We excluded reviews and papers not containg case description as well as cases which documented an appropriate response to injected adrenalin.
+Also cases describing Mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
+The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
+
+We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment.
+We used piloted polls to gather the data and all reports were screened once by the same researcher.
+
+All of the studies were case reports therfore the evidence level was classified as low.
+
+The review protocol was not established prior to the review and the review did not undergo preregistration.
+
+We have performed literature research identifying case reports of refractory anaphylaxis.
+All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included into this study.
+We investigated which interventions were reported as succesfull in the treatment of anaphylactic hypotension that was not reponding to injected (i.v.
+or i.m) adrenailn.
+Study outcomes were defined as time to elicit a response and mortality.
+Most of these reports described refractory anaphylaxis in a perioperative setting, but they also included cases of food and drug anaphylaxis.
+Suprisingly, no cases of insect venom anphylaxis were found among these reports.
+
+Outcomes: successful and unsuccesful managment of anaphylaxis was one of the most important outcomes we evaluated.
+This information was availible in all cases.
+Neurological deficits directly after the episode were also evaluated, but this measure might be biased.
+
+<!-- ## Data items -->
+<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->
+
+<!-- ## Risk of bias in individual studies -->
+<!-- Describe methods used for assessing risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis.
+-->
+
+<!-- ## Summary measures -->
+
+<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->
+
+<!-- ## Synthesis of results -->
+
+<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->
+
+<!-- ## Risk of bias across studies  -->
+
+<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->
+
+<!-- ## Additional analyses  -->
+
+<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->
+
+```{r demograph}
+source("../../R/data_handling.R")
+knitr::kable(demogr.tab,col.names = c("Elicitor","Age","Male fraction","Operative fraction"),digits = 1)
 ```
 
-# Methods
+# Risk factors of refractory anaphylaxis
 
-# Results
+The risk factors for refractory anaphylaxis include perioperative anaphylaxis. We observed XXX cases. 
+Fatality rate of anaphylaxis during anaesthesia is estimated in about 3-10% of cases [@Schummer2008].
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
+Asthma poses higher risk of anaphylaxis [@Koplin2011] but we have not noted any cases of asthma in the reporter refractory anaphylaxis case reports.
+
+The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anaesthesia drugs, aprotinin and protamine, radiocontrast media (RCM).
+
+While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000---20,000 anesthesia pro- cedures, it is responsible for 3---10% of the perioperative fatalities. [@Nag2017]
+
+<!-- # Firs sings in anesthesia setting of anaphylaxis -->
+1. In our two cases ETco2 collapse was the first sign preceding cardiac arrest. Several authors have underlined
+the value of this early clinical sign for the
+suspicion of anaphylactic shock, even in the absence of
+any cutaneous manifestations. [@Baumann2009]
+
+# Therapy of refractory anaphylaxis
+
+Most of the treatment strategies were according to the guidelines of anaphylaxis management [source], but were supplemented by additional procedures that saved patients' lives.
+We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
+
+Ambulatory patients and not in hospital-setting were first treated with adrenaline wether patients experiencing anaphylaxis under anesthesia were usually first given intraveous volume replacement and 100% oxygen, followed by eithe epinephrine or alfa-adrenergic vasopressors or both.
+
+Epinephrine was given in 100% of cases in various doses.
+
+In Table X we provided therapeutic approaches used in the treatment of anaphylaxis.
+
+4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.
+
+```{r tabFatal}
+knitr::kable(fat)
 ```
 
-# Discussion
+*The AAAAI/ACAAI Guidelines show info about refractory anaphylaxis [@Simons2014] The three guidelines on the management of refractory anaphylaxis: [@Simons2014]  Management of refractory anaphylaxis WAO: intubation; ventilation; IV vasopressors; glucagon; anticholinergic; transfer to hospital (preferably to an emergency medicine, critical care medicine, or anesthesiology) team for ventilatory and inotropic support; checklist of needed items; Table 6 AAAAI: vasopressors; dopamine; give vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension; transfer to hospital; glucagon; atropine; methylene blue; includes checklist of supplies and equipment; Figures E2, E3 EAACI: glucagon*
+
+*Management of anaphylaxis refractory to initial treatment The guidelines differ in emphasis on refractory anaphylaxis treatment.
+The WAO Guidelines stress the importance of prompt initial treatment to prevent escalation of symptoms [2].
+They suggest that if possible, patients with anaphylaxis refractory to epinephrine, supplemental oxygen, IV fluids, and second-line medications should be transferred to the care of a specialist team for ventilatory and inotropic support and continuous electronic monitoring [2].
+The AAAAI/ACAAI Guidelines provide details about interven- tions for cardiopulmonary arrest, airway management, and IV administration of vasopressors including epinephrine, dopamine, and vasopressin [3].
+The EAACI Guidelines in- clude brief specific instructions about when to call for In- tensiveCareUnit support [4](Table4).
+Studies relevant to refractory anaphylaxis treatment are of interest.
+A Cochrane review of randomized con- trolled trials (RCT) in more than 20,000 critically ill pa- tients with distributive shock supports administration of crystalloids such as 0.9% saline, because administration of colloids such as albumin or hetastarch did not correlate with increased survival [82].
+Methylene blue administration for vasoplegia in anaphylaxis refractory to epinephrine and IV fluid resuscitation is based on case reports and extrapolation from use in other forms of shock [83].
+[@Simons2014]*
+
+*Management of anaphylaxis refractory to initial treatment Even in high-resource countries, optimal treatment of re- fractory anaphylaxis is not available universally; for ex- ample, in remote, inaccessible, or impoverished areas or in specific situations such as anaphylaxis on airplanes.
+In limited-resource situations, lack of availability of basic es- sentials such as epinephrine, supplemental oxygen and IV fluid resuscitation is more critical than lack of second-line medications such as antihistamines and glucocorticoids.
+Lack of availability of advanced life-support management can be a major barrier to survival [72-76,128] (Table 8).
+In any limited-resource situation, resuscitation efforts pro- longed over hours using a hand-held bag valve mask (manual resuscitator) are often successful in anaphylaxis [2] (Table 8).
+In mid- and low-resource countries, striving to ensure more consistent availability of medications, supplies, and equipment for anaphylaxis treatment is an important goal [2,5-7].
+The World Health Organization has devel- oped a tool kit containing evidence-based guidelines and  a framework for quality improvement in the hospital care of critically ill children in such environments [126], where despite many obstacles, improvements can be documented [130].
+[@Simons2014]*
+
+*No high-quality studies at present permit recommendations concerning the use of intravenous epinephrine in anaphylaxis [18] Parameters and international guidelines support its use in refractory anaphylaxis [20–23] (Table 1).
+BUt it can cause arrythmias that are lethal.
+[@Kemp2014] *
+
+# Methylene blue for the treatment of refractory hypotension
+
+The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce Nitric oxide (NO).
+NO subsequently activates the soluble guanyl cyclase (sGC) which is responsible for the producion of cGMP out of guanosine triphosphate (GTP) [@Lucas2000].
+This in effect leads to the opening of the calcium ion channels in smooth muscle tissues and subsequent loss of vasal muscle contractility [@Omar2015].
+
+Methylene blue is a well investigated drug used for treating patients with methemoglobinemia [@Schirmer2003].
+Buzato et al.
+investigated its properties as an inhibitor of nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue was mediated by the sGC inhibition.
+Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].
+
+`r which(d$What.helped_gr=="methylene blue") %>% length` reports (`r ((which(d$What.helped_gr=="methylene blue") %>% length)/length(d$What.helped_gr)*100) %>% round(1)`%) repoted using methylene blue to treat refractory anaphylaxis and indicated a rapid (within minutes) increase in blood pressure after the injection of methylene blue.
+It was both given as a bolus and as an infusion with the doses ranging from 0.5-2mg/kg.
+Patients responded well to the treatment with methylene blue and a second infusion (or bolus) were successfully given in case of a recurrent hypotension within hours of the anaphylactic episode.
+
+Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis withouth hypotension [@Bauer2013].
+Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) [7-9] production which is a potent mediator of anaphylaxis [].
+
+The use of methylene blue in current managment of anaphylaxis is scarce.
+We recieved no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.
+
+We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension using 100 mg bolus i.v.
+and follow it with an infusion of 50 µg/kg/min.
+As a vector a 100 mL of 5% dextrose can be used.
+
+As methylene blue does not block nitric oxide production its side effects are minimal (Nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
+It may be considered a safe option for the treatment of anaphylaxis.
+No side effects were reported due to methylene blue use in the above cases.
+
+It is important to remember that methylene blue may provoke false pulseoximetry readings as the blue dye mimics cyanosis [@Kessler1986].
+
+# Vasopressors Vasopressin as a supplementary vasopressor
+
+In case of a hypotension refractory to adrenaline or volume replacement therapy use of vasopressors can be evaluated.
+
+Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
+Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases.
+In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes.
+Other vasopressors included metaraminol, methoxamine, dopamine and noradrenaline infusions.
+
+Interantional guidelines support the use of dopamine in refractory hypotension [@Simons].
+
+Our register lists `r ` 18 cases where dopamine was used. None were fatal.
+
+*The authors state that “current recommendations are to start with an infusion of epinephrine for unresponsive anaphylaxis; and if there is re- fractory hypotension, then add dopamine as a second vaso- pressor.”1 The literature supports the use of epinephrine infusion in refractory anaphylactic shock because it reverses most of the cardiovascular effects of anaphylaxis.3 At low doses, dopamine exerts its effects on the b-adrenergic receptors, causing an in- crease in myocardial function, heart rate, and some vasodilata- tion.
+However, there is no clear evidence for the use of dopamine as a second-line vasopressor in anaphylactic shock.
+This is particularly important given that dopamine has been associated with a higher risk of mortality and major adverse events compared with epinephrine and norepinephrine.4,5 The literature suggests that norepinephrine or vasopressin may be appropriate second-line agents.
+Norepinephrine acts primarily on the a-adrenergic receptors, whereas vasopressin acts on the V1 re- ceptors, both leading to vasoconstriction.2,6 Methylene blue has also been used with good success in catecholamine-refractory vasoplegic anaphylactic shock.7 [@Alquarashi2016]*
+
+Seedat et al described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anaesthesia.
+Reafractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone.
+Rapid response was achieved as a last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occured.
+Authors concluded that vasopressin or ornipressin should be considered for the treatment
+of adrenaline-resistant anaphylactic shock before resuscitation is discontinued. [@Seedat2014]
+
+# Glucagon in anaphylaxis with concomitant beta blockers
+
+*There is evidence that glucagon is also helpful in refractory anaphylaxis especially for patients on beta-blockers.
+Glucagon activates adenylyl cyclase directly and can bypass the beta blockade.11,12 [@Coop2017]*
+
+*Histamine release also increases cardiac rate, cardiac contractility, and bronchoconstriction.2,5 Therefore, it is suggested that beta-blockers mask cardiac signs of anaphylaxis and lead to unopposed alpha-adrenergic activity, causing severe bronchoconstriction.
+[@Coop2017]*
+
+*Finally, we would like to suggest a correction in the published glucagon dose.
+In children, the loading dose is 20 to 30 µg/kg (maximum, 1 mg), and the infusion dose is 5 to 15 µg/min rather than the dose recommended in the 2015 update [@Alquarashi2016]*
+
+*Bradycardia is an uncommon presentation during anaphylaxis. In our patient, premedication with beta blocker and severe hypoxia could be the probable reason for this uncommon presentation.[@Nag2017]*
+
+# ECLS
+*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*
+
+*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*
+
+*Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
+10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - return of spontaneaous circulation.
+[@Wang2016] This case shows how important it is to get rid of the alergen in refractory anaphylaxis.
+Traetment with ECMO.*
+
+# Stopping the elicitor exposure
+In cases where the elicitor of anaphylaxis is known, and first line  therapy fails it is advisable to stop the exposition to the likely allergen.
+Three case reports described rapid improvement of a refractory anaphylaxis after preventing further contact with the alergen.
+
+A 15-year-old boy who whas exposed to cow's milk underwent near fatal anaphylaxis with refractory hypotension.
+After 150 minutes of resuscitative measures a gastric drainage performed as a last resort treatment suprisingle led to a quick alleviation of anaphylaxis symptoms and patient recovered fully.
+Authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].
+
+Another report used sugammadex which is a compund strongly binding to rocuronium, to decrease its bioavailiblity in a patient under anaesthesia and therefore allow for cesation of a refractory anaphylaxis episode [@Raft2012].
+
+There also was a report where patient was inadvertently given i.v.
+high dose of aprotinin in a fast infusion.
+After 10 hours of refractory hypotension authors decided to use the last resort treatment with high-volume continuous venovenous hemofiltration (HV-CVVH) which resulted in a rapid improvement of hemodynamic function [@Lango2009].
 
 # Conclusion 
+<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article.
+Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->
 
-# Acknowledgements
+# Expert Commentary: 500-1000 words (included in overall word count).
+<!-- **To distinguish the articles published in the Expert Review series, authors must provide an additional section entitled ‘Expert Commentary’.
+This section affords authors the opportunity to provide their interpretation of the data presented in the article and discuss the developments that are likely to be important in the future, and the avenues of research likely to become exciting as further studies yield more detailed results.
+The intention is to go beyond a conclusion and should not simply summarise the paper.
+Authors should answer the following:** -->
+
+## What are the key weaknesses in clinical management so far?
+
+## What potential does further research hold? What is the ultimate goal in this field?
+
+## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?
+
+## Is there any particular area of the research you are finding of interest at present?
+<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->
+
+# Five-year view
+<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->
+
+# Key issues
+<!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->
+
+<!-- * Matrix-Assisted Laser Desorption/Ionization  Mass Spectrometry Imaging (MALDI-MSI) is able to provide a molecular insight of the samples, detecting the presence of a great variety of analytes directly on-tissue and showing their spatial distribution across the tissue section.
+-->
+
+*?
+Epinephrine in appropriate doses is safe, and there are no abso- lute contraindications for its use in treating anaphylaxis.
+?
+Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.*[@Fineman2015]
+* Glucagon
+* Vasopressors
+* ECLS
+*
 
 # References
+<div id="refs"></div>
+<!-- **A maximum of 100 references is suggested.
+Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography.
+Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->
 
-### Colophon
+## Annotated bibliography
+<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+# Figures and Tables
+<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->
+
+```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
+source("../../R/prisma.R")
+knitr::include_graphics("../figures/prism_flow.png")
+```
+
+<!-- ### Colophon -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
 
@@
 # which R packages and versions?
-devtools::session_info()
+# devtools::session_info()
 ```
 
-The current Git commit details are:
+<!-- The current Git commit details are: -->
 

### 2017-11-23 R/analysis.R

@@ -23,15 +23,34 @@ guineapigs<- c(141,149) %>% length()
 no_case_rep <- 131 %>% length()
 
 # Include information from the anaphylaxis registry
-#load("~/Documents/anaphylaxis/data3.R")
+load("~/Documents/anaphylaxis/data3.R")
 ## See multiple adrenalin cases
-#require(magrittr)
-#data3$d_560_adren2_v5 %>% summary
+require(magrittr)
+data3$d_560_adren2_v5 %>% summary
 #data3[data3$d_560_adren2_v5=="yes",15:51] %>% summary
 #data3$q_142_fatal_treatment_v5%>% summary
 #data3[data3$q_142_fatal_treatment_v5=="yes",51:100] %>% summary
 
 
 
+### Here are some results from free treatment field of our registry - these indicate refractory anahpylaxis.
+insgesamt 3 Adrenalin-injektionen durch Hausarzt
 
+Ephedrin 50mg, Infusionen,Adrenalin als wiederholter Bolus (100 mikrogramm)ok, dann Adrenalinperfusor, Noradrenalinperfusor, Trendelenburglagerung
+
+Ephedrin 40mg, Magnesium 2 Ampullen, Defibrillation 3x 200Joule, Amiodaron 1 Ampulle
+
+Atropin mehrfach 0,5 mg
+Atropin Boli, 2 minütige mechanische und medikamentöse Reanimation
+Reanimation, weitere Medi unbekannt
+
+# We had two glucagon treated anaphylaxises here !!!
+data3$q_522_glucagon_v5 %>% summary()
+data3[which(data3$q_522_glucagon_v5=="yes"),200:300]
+
+
+# Our registry lists cases with dopamine:
+dop <- (data3$q_522_dopamine == "yes") %>% which()
+names(data3)
+data3[dop,c(11,12,66,74,366,297,43,44,45,46,47,48,49)]
 

### 2017-11-23 R/data_handling.R

@@ -1,5 +1,6 @@
 ######## Load the data ########
 d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
+d <- read.csv("~/Documents/refractoryANA/analysis/data/raw_data/refractoryAnaphylaxis.csv")
 
 ######## prepearing the data frame ######
 mean(d$age,na.rm = T)
@@ -141,7 +142,25 @@ d$t_antihistamines <- c(T,T,T,F,T,F,F,F,F,T,
                         F,F,T,F,F,F,F,F,F,T,
                         T,F,F,T,F,F,F,F,T)
 
-# d$t_volume <- c()
+d$t_vasopressin <- rep(F,length(d$Therapy))
+d$t_vasopressin[c(3,10,11,27,28,29,30,31,32,33,35)] <- T
+terlipressin <- c(14,15)
+d$t_vasopressin[terlipressin] <- T
+d$t_vasopression <-  d$t_vasopressin
+metaraminol <-c(1, 6,16,17)
+isoprenaline.dobutamine <- c(22 ,36)
+norepi <- c(27,28,30,31,32,33,35,36,1,2,3,5,10)
+d$t_vasopression[c(metaraminol,isoprenaline.dobutamine,norepi)] <- T
+
+
+cardiac_bypass <- c(10,11,15,3,6,19)
+
+ALS <- c(17,18,19,21,25,30,39,6)
+
+b2 <- c(12,2,3,4)
+
+glucagon <- c(20,24)
+
 
 ###### Fatalities table ######
 

### 2017-11-23 R/refracrory_graph.R

@@ -1,0 +1,67 @@
+# Refractory ANA flow diagdramm
+require(DiagrammeR)
+
+create_graph() %>%
+  add_node(x = 0, y = 0,
+           label="Anaphylaxis suspected:\n1. Assess ABC\n2. Inject 300 µg adrenalin i.m.\n3. Call for help",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 0, y = -1,
+           label="1. Venous access\n2. 100% oxygen\n3. Crystalloids i.v\n4. Second dose of adrenalin\n5. H1+H2 Antihistamines i.v.\n6. 125 mg Prednisone i.v.",
+           shape="rectangle",
+           width=2.2,
+           height=1) %>%
+  add_node(x = -2, y = -2,
+           label="Concomitant beta-blockers?",
+           shape="oval",
+           width=2.5) %>%
+  add_node(x = 0, y = -2, label="Refractory bronchospasm?",
+           shape="oval",
+           width=2) %>%
+  add_node(x = 2, y = -2, label="Refractory hypotension?",
+           shape="oval",
+           width=2) %>%
+  add_node(x = 4, y = -1, label="1. Observation (biphasic reaction)\n2. Adrenalin auto-injector\n3. Allergology consult",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = -2, y = -3, label="consider glucagon infusion",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 2, y = -3, label="consider vasopressors (dopamine, vasopressine)",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 0, y = -3, label="consider inhalative beta2-mimetics",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 4, y = -2, label="Known allergen?",
+           shape="oval",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 4, y = -3, label="consider gastric drainage (food)\nsugammadex (rocuronium)\nHaemofiltration (drug)",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_edge(1,2) %>%
+  add_edge(1,6) %>%
+  add_edge(2,3) %>%
+  add_edge(2,4) %>%
+  add_edge(2,5) %>%
[…]

### 2017-11-23 analysis/paper/paper.Rmd
grammarly checked

--- 2017-11-20 analysis/paper/paper.Rmd
@@
 library(refractoryANA)
+require(magrittr,quietly = T)
+
 ```
+
+Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+
+**Corresponding author: **
+Prof. Dr. med. M. Worm margitta.worm@charite.de
+Phone: +49 30 450 529 005;      Fax: +49 30 450 529 902
 
@@
 
-Here is a citation [@Marwick2017]
+Anaphylaxis is a life-threatening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties.
+Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or respiratory compromise, or when the allergen is known or likely [@Sampson2006].
+Its pathogenesis relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-# Background
+Anaphylaxis remains a clinical diagnosis [@Kemp2014], but consensus criteria of anaphylaxis suggest a standardized set of criteria [@Sampson2006].
+The definition of _refractory_ anaphylaxis is not established.
+For the purpose of this review, we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
+Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e.
+rapid increase in blood pressure, bronchodilatation, tachycardia).
 
-```{r demo, eval = FALSE}
-1 + 1
+It is important to distinguish refractory anaphylaxis from recurrent anaphylaxis.
+In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
+We did not include such reports in our analysis and advise on using the term "recurrent" anaphylaxis instead of "refractory" in such cases.
+
+## Epidemiology
+
+The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
+It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015].
+Anaphylaxis in children is usually food dependent and less severe than episodes in adulthood.
+Cases of _refractory_ anaphylaxis are extremely rare.
+The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.
+
+## Risk factors for anaphylaxis
+
+Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013].
+Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigorous exercise (e.g.
+jogging) [@Worm2013].
+Risk factors for _refractory_ anaphylaxis were not investigated previously.
+
+## Management
+
+The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.
+The mainstay of therapy is the early administration of intramuscular (i.m.) adrenalin [@Lazar2017].
+However, optimal anaphylaxis treatment is difficult to study because it is an emergency condition and placebo-controlled studies are impossible.
+Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.
+
+We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
+
+<!-- We compared the successful procedures from the case reports to ??? -->
+
+<!-- ## Eligibility criteria -->
+<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale.
+-->
+
+<!-- ## Information sources -->
+<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->
+
+## Literature search strategy
+
+We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R].
+The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017.
+The search was conducted on 16.11.2017 and returned 150 results.
+All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included in this study.
+Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded.
+Additionally, case reports referenced in the included sources were added.
+We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenalin.
+Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
+We finally included 29 Papers reporting ` r d$Author %>% length()` cases of anaphylaxis.
+The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
+
+We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment.
+We used piloted polls to gather the data and all reports were screened once by the same researcher.
+We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v.
+or i.m.) adrenalin.
+Study outcomes were defined as time to elicit a response and mortality.
+Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated.
+This information was available in all cases.
+All of the studies were case reports therefore the evidence level was classified as low.
+The review protocol was not established prior to the review and the review did not undergo preregistration.
+
+<!-- ## Data items -->
+<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->
+
+<!-- ## Risk of bias in individual studies -->
+<!-- Describe methods used for assessing the risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis.
+-->
+
+<!-- ## Summary measures -->
+
+<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->
+
+<!-- ## Synthesis of results -->
+
+<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->
+
+<!-- ## Risk of bias across studies  -->
+
+<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->
+
+<!-- ## Additional analyses  -->
+
+<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->
+
+```{r demograph}
+source("../../R/data_handling.R")
+knitr::kable(demogr.tab,col.names = c("Elicitor","n","Age","Male fraction","Operative fraction"),digits = 1)
 ```
 
-# Methods
+# Risk factors for refractory anaphylaxis
 
-# Results
+The risk factors for refractory anaphylaxis include perioperative anaphylaxis.
+In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the reaction happened during a diagnostic or surgical procedure.
+The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM).
+Surprisingly, no cases of insect venom anaphylaxis were found among these reports.cWhile the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities.
+[@Nag2017]
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
+<!-- # Firs sings in anesthesia setting of anaphylaxis -->
+Early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
+[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.
+
+# Therapy of refractory anaphylaxis
+
+Most of the treatment strategies were according to the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed.
+We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
+
+Based on the analyzed reports, ambulatory patients were first treated with adrenaline were patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, followed by steroid and antihistaminic agents.
+Epinephrine was given in 100% of cases in various doses.
+
+<!-- In Table X we provided therapeutic approaches used in the treatment of anaphylaxis.  -->
+
+4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.
+
+```{r tabFatal}
+rownames(fat) <- NULL
+colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
+knitr::kable(fat)
 ```
 
-# Discussion
+World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011].
+Patients should be transported to a critical care unit.
+AAAAI [@Lieberman] suggest similar treatment but include dopamine, vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension, glucagon atropine, and methylene blue.
+EAACI [@Muraro2014] guidelines also suggest using glucagon in case of concomitant beta-blockade.
+
+# Methylene blue for the treatment of refractory hypotension
+
+`r which(d$What.helped_gr=="methylene blue") %>% length` reports (`r ((which(d$What.helped_gr=="methylene blue") %>% length)/length(d$What.helped_gr)*100) %>% round(1)`%) reported using methylene blue to treat refractory anaphylaxis and indicated a rapid (within minutes) increase in blood pressure after the injection of methylene blue.
+It was both given as a bolus and as an infusion with the doses ranging from 0.5-2mg/kg.
+Patients responded well to the treatment with methylene blue and a second infusion (or bolus) were successfully given in case of a recurrent hypotension within hours of the anaphylactic episode.
+
+The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce Nitric oxide (NO).
+NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000].
+This in effect leads to the opening of the calcium ion channels in smooth muscle tissues and subsequent loss of vascular muscle contractility [@Omar2015].
+
+Methylene blue is a well-investigated drug used for treating patients with methemoglobinemia [@Schirmer2003].
+Buzato et al.
+investigated its properties as an inhibitor of nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue was mediated by the sGC inhibition.
+Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].
+
+Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013].
+Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) *[7-9]* production which is a potent mediator of anaphylaxis [].
+
+The use of methylene blue in current management of anaphylaxis is scarce.
+We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.
+
+We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension using 100 mg bolus i.v.
+and follow it with an infusion of 50 µg/kg/min.
+As a vector, a 100 mL of 5% dextrose can be used.
+
+As methylene blue does not block nitric oxide production its side effects are minimal (Nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
+It may be considered a safe option for the treatment of anaphylaxis.
+No side effects were reported due to methylene blue use in the above cases.
+
+It is important to remember that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].
+
+# Vasopressors Vasopressin as a supplementary vasopressor
+
+In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated.
+
+Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
+Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases.
+In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes.
+Other vasopressors included metaraminol, methoxamine, dopamine and noradrenaline infusions.
+
+Our register lists 18 cases where dopamine was used. None were fatal.
+
+Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Liberman] and was never vastly investigated.
+Dopamine, however, is recommended by 2 out of three guidelines.
+Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni].
+Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
+Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@Chiara2008].
+
+Seedat et al.
+described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia.
+The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone.
+Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred.
+Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued.
+[@Seedat2014] We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.
+
+One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell].
+Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.
+
+# Glucagon in anaphylaxis with concomitant beta blockers
+
+Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986, @Javeed1996].
+Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
+
+Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis. [@Nag2017]*
+
+# ECLS
+
+`r (d$What.helped_gr=="ECLS") %>% which()`
+*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*
+
+*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*
+
+*Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
+10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - the return of spontaneous circulation.
+[@Wang2016] This case shows how important it is to get rid of the allergen in refractory anaphylaxis.
+Treatment with ECMO.*
+
+# Stopping the elicitor exposure
+In cases where the elicitor of anaphylaxis is known, and first-line  therapy fails it is advisable to stop the exposition to the likely allergen.
+Three case reports described the rapid improvement of a refractory anaphylaxis after preventing further contact with the allergen.
+
+A 15-year-old boy who was exposed to cow's milk underwent near-fatal anaphylaxis with refractory hypotension.
+After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylaxis symptoms and patient recovered fully.
+Authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].
+
+Another report used sugammadex which is a compound strongly binding to rocuronium, to decrease its bioavailability in a patient under anesthesia and therefore allow for a cessation of a refractory anaphylaxis episode [@Raft2012].
+
+There also was a report where the patient was inadvertently given i.v.
+high dose of aprotinin in a fast infusion.
+After 10 hours of refractory hypotension, authors decided to use the last resort treatment with high-volume continuous venovenous hemofiltration (HV-CVVH) which resulted in a rapid improvement of hemodynamic function [@Lango2009].
 
 # Conclusion 
+<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article.
+Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->
 
-# Acknowledgements
+# Expert Commentary: 500-1000 words (included in overall word count).
+<!-- **To distinguish the articles published in the Expert Review series, authors must provide an additional section entitled ‘Expert Commentary’.
+This section affords authors the opportunity to provide their interpretation of the data presented in the article and discuss the developments that are likely to be important in the future, and the avenues of research likely to become exciting as further studies yield more detailed results.
+The intention is to go beyond a conclusion and should not simply summarise the paper.
+Authors should answer the following:** -->
+
+## What are the key weaknesses in clinical management so far?
+
+## What potential does further research hold? What is the ultimate goal in this field?
+
+## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?
+
+## Is there any particular area of the research you are finding of interest at present?
+<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->
+
+# Five-year view
+<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->
+
+# Key issues
+<!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->
+
+<!-- * Matrix-Assisted Laser Desorption/Ionization  Mass Spectrometry Imaging (MALDI-MSI) is able to provide a molecular insight of the samples, detecting the presence of a great variety of analytes directly on-tissue and showing their spatial distribution across the tissue section.
+-->
+
+*?
+Epinephrine in appropriate doses is safe, and there are no abso- lute contraindications for its use in treating anaphylaxis.
+?
+Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.*[@Fineman2015]
+* Glucagon
+* Vasopressors
+* ECLS
+*
 
 # References
+<div id="refs"></div>
+<!-- **A maximum of 100 references is suggested.
+Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography.
+Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->
 
-### Colophon
+## Annotated bibliography
+<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+# Figures and Tables
+<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->
+
+```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
+source("../../R/prisma.R")
+knitr::include_graphics("../figures/prism_flow.png")
+```
+
+<!-- ### Colophon -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
 

### 2017-11-23 R/data_handling.R

@@ -1,14 +1,17 @@
 ######## Load the data ########
-d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
+# d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
 d <- read.csv("~/Documents/refractoryANA/analysis/data/raw_data/refractoryAnaphylaxis.csv")
 
 ######## prepearing the data frame ######
 mean(d$age,na.rm = T)
 d$Elicitor_gr <- rep(NA,length(d$Elicitor))
-d$Elicitor_gr[c(grepl("cont",d$Elicitor),24)] <- "contrast"
+d$Elicitor_gr[c(which(grepl("cont",d$Elicitor)==T),24)] <- "contrast"
 d$Elicitor_gr[c(4,7,13,19,20,21,38,39)] <- "unknown"
-d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,23,27,28,29,30,31,32,33,34,35,36,40,41)] <- "drug"
+d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,23,27,28,29,30,31,32,33,34,35,36,40,41,43)] <- "drug"
 d$Elicitor_gr[c(1,3,5,15,25,37,42)] <- "other"
+
+# d$Elicitor[21] <- "platinum"
+d$Elicitor[22] <- "contrast"
 
 d$Time.after.exposure
 d$Time <- c(5,5,10,NA,5,NA,5,NA,NA,NA,NA,NA,NA,NA,15,NA,NA,5,5,NA,NA,5,5,0.2,5,5,0.5,NA,5,20,5,5,10,5,5,5,15,NA,NA,0.5,0.5,5,NA)
@@ -78,7 +81,8 @@ demogr.tab <- data.frame(Age=d$age %>% split(d$Elicitor_gr) %>%
              lapply(function(x){
                which(x=="OP")  %>% length()/length(x)*100
              }) %>% unlist
-) %>% data.frame(Elicitor=rownames(.),.)
+) %>% data.frame(Elicitor=rownames(.),
+                 n=d$Elicitor_gr %>% factor %>% summary() %>% unlist,.)
 
 rownames(demogr.tab) <- NULL
 
@@ -130,7 +134,7 @@ d$What.helped_gr <- c("methylene blue",
 
 ##### Therapy dissection ####
 
-d$Therapy
+# d$Therapy
 
[…]

### 2017-11-23 analysis/paper/paper.Rmd
corrected the prisma and references missing

@@
 
-Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Liberman] and was never vastly investigated.
+Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Lieberman] and was never vastly investigated.
 Dopamine, however, is recommended by 2 out of three guidelines.
-Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni].
+Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni2015].
 Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
-Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@Chiara2008].
+Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].
 
@@
 
-One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell].
+One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
 Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.

### 2017-11-23 R/prisma.R

@@ -24,11 +24,11 @@ create_graph() %>%
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 3\nnot a case report = 3\nnot anaphylaxis = 2\nanimal study = 2\njapanese language = 1",
+  add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 6\nnot a case report = 4\nnot anaphylaxis = 3\nanimal study = 3\njapanese language = 1",
            shape="rectangle",
            width=2,
            height=1) %>%
-  add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 37)",
+  add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 29)",
            shape="rectangle",
            width=2,
            height=0.6) %>%

### 2017-12-04 analysis/paper/paper.Rmd
references update, Glucagon WIP

@@
   - Wojciech Francuzik
+  - Sabine Dölle
   - Margitta Worm, MD
@@
 
-```{r, setup, echo = FALSE}
+```{r, setup, echo = FALSE,include=F}
 knitr::opts_chunk$set(
@@
 
-library(refractoryANA)
+#library(refractoryANA)
 require(magrittr,quietly = T)
+# source("../../R/data_handling.R")
+devtools::load_all(".",quiet = T)
 
@@
 # Introduction
+Anaphylaxis is a potentially life-threatening mast cell dependent reaction to  most frequently food, drugs, insect venom.
+Anaphylaxis is a clinical diagnosis.
+It is highly likely if there is an acute onset of skin symptoms in conjuction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
+The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-Anaphylaxis is a life-threatening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties.
-Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or respiratory compromise, or when the allergen is known or likely [@Sampson2006].
-Its pathogenesis relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
-
-Anaphylaxis remains a clinical diagnosis [@Kemp2014], but consensus criteria of anaphylaxis suggest a standardized set of criteria [@Sampson2006].
-The definition of _refractory_ anaphylaxis is not established.
-For the purpose of this review, we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
+Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast the term refractory anaphylaxis has not established.
+For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
 Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e.
@@
 In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
-We did not include such reports in our analysis and advise on using the term "recurrent" anaphylaxis instead of "refractory" in such cases.
+We did not include such reports into our analysis and suggest to use the term “recurrent” anaphylaxis instead of “refractory” in such cases.
 
@@
 It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015].
-Anaphylaxis in children is usually food dependent and less severe than episodes in adulthood.
-Cases of _refractory_ anaphylaxis are extremely rare.
-The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.
+Anaphylaxis in children is in the majority of cases food dependent and most likely more frequently less severe than episodes in adulthood (Ref).
+Cases of refractory anaphylaxis are extremely rare.
+The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3-5% of all cases of anaphylaxis.
 
@@
 The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.
-The mainstay of therapy is the early administration of intramuscular (i.m.) adrenalin [@Lazar2017].
-However, optimal anaphylaxis treatment is difficult to study because it is an emergency condition and placebo-controlled studies are impossible.
-Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.
+Until to day the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
+However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are for ethical reasons limited.
+Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.
 
-We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
+In the present analysis we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
 
@@
 The search was conducted on 16.11.2017 and returned 150 results.
-All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included in this study.
+All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study.
 Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded.
 Additionally, case reports referenced in the included sources were added.
-We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenalin.
+We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenaline.
 Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
-We finally included 29 Papers reporting ` r d$Author %>% length()` cases of anaphylaxis.
+We finally included 29 Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
 The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
@@
 We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v.
-or i.m.) adrenalin.
+or i.m.) adrenaline.
 Study outcomes were defined as time to elicit a response and mortality.
@@
 ```{r demograph}
-source("../../R/data_handling.R")
-knitr::kable(demogr.tab,col.names = c("Elicitor","n","Age","Male fraction","Operative fraction"),digits = 1)
+knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1)
 ```
@@
 
-The risk factors for refractory anaphylaxis include perioperative anaphylaxis.
-In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the reaction happened during a diagnostic or surgical procedure.
+A major risk factor for refractory anaphylaxis is perioperative anaphylaxis.
+In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occured during a diagnostic or surgical procedure.
 The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM).
-Surprisingly, no cases of insect venom anaphylaxis were found among these reports.cWhile the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities.
+Surprisingly, no cases of insect venom anaphylaxis were found among these reports.
+While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 – 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 – 10% of the perioperative fatalities.
 [@Nag2017]
@@
 <!-- # Firs sings in anesthesia setting of anaphylaxis -->
-Early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
+An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
 [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.
@@
 Patients should be transported to a critical care unit.
-AAAAI [@Lieberman] suggest similar treatment but include dopamine, vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension, glucagon atropine, and methylene blue.
-EAACI [@Muraro2014] guidelines also suggest using glucagon in case of concomitant beta-blockade.
+The AAAAI [@Lieberman] is  suggesting a similar treatment but include dopamine and vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension.
+Also the EAACI [@Muraro2014] guidelines suggest to use glucagon in case of concomitant beta-blockade.
 
@@
 Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013].
-Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) *[7-9]* production which is a potent mediator of anaphylaxis [].
+Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) production which is a potent mediator of anaphylaxis [@Moritoki1992].
 
@@
 
-As methylene blue does not block nitric oxide production its side effects are minimal (Nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
+As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
 It may be considered a safe option for the treatment of anaphylaxis.
@@
 
-It is important to remember that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].
+Clinicians must be aware that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].
 
-# Vasopressors Vasopressin as a supplementary vasopressor
+# Vasopressors
 
@@
 In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes.
-Other vasopressors included metaraminol, methoxamine, dopamine and noradrenaline infusions.
+Other vasopressors included metaraminol 
+[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
+[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
+[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
+Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug.
+One patient rected to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].
 
-Our register lists 18 cases where dopamine was used. None were fatal.
-
-Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Lieberman] and was never vastly investigated.
+Use of vasopressin in refractory anaphylaxis was only suggested in  management recommendations from the AAAAI [@Lieberman] and was never vastly investigated.
 Dopamine, however, is recommended by 2 out of three guidelines.
-Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni2015].
+Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015].
 Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
@@
 described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia.
-The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone.
+The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone [@Seedat2014].
 Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred.
-Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued.
-[@Seedat2014] We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.
+Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014].
+We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.
 
@@
 
+The European anaphylaxis register lists 18 cases where dopamine was used.
+None of these reactions were fatal and  Ring and Messmer.
+`r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure.
+Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
+
 # Glucagon in anaphylaxis with concomitant beta blockers
 
-Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986, @Javeed1996].
+Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Patients
+
 Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
@@
 Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis. [@Nag2017]*
-
-# ECLS
-
-`r (d$What.helped_gr=="ECLS") %>% which()`
-*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*
-
-*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*
-
-*Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
-10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - the return of spontaneous circulation.
-[@Wang2016] This case shows how important it is to get rid of the allergen in refractory anaphylaxis.
-Treatment with ECMO.*
 
@@
 
+# ECLS
+
+`r (d$What.helped_gr=="ECLS") %>% which()`
+*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*
+
+*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*
+
+*Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
+10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - the return of spontaneous circulation.
+[@Wang2016] This case shows how important it is to get rid of the allergen in refractory anaphylaxis.
+Treatment with ECMO.*
+
 # Conclusion 
@@
 
-## What are the key weaknesses in clinical management so far?
+Severe anaphylaxis is a critical medical condition requiring an immediate intervention.
+According to the international guidelines adrenaline intramuscular given is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
+However, in a few patients this intervention might not be sufficient and besides repeating the doses other therapeutical measures will need to be applied.
+Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously.
+This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the ….clotting …system??
+check.
+In clinical practice it is necessary to identify patients of risk.
+Such risk factors should particulary be assessed in patients undergoing medical diagnostic and therapeutic measures.
+The measurement of tryptase enables to identify patients how are at risk for severe anaphylaxis.
 
-## What potential does further research hold? What is the ultimate goal in this field?
-
-## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?
-
-## Is there any particular area of the research you are finding of interest at present?
-<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->
-
-# Five-year view
-<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->
+For the future, biomarker measurements in a bed side setting may enable doctors to identify  patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
+Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can safe life in a given patient.
 
@@
 ```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
-source("../../R/prisma.R")
+#source("../../R/prisma.R")
 knitr::include_graphics("../figures/prism_flow.png")
+```
+
+```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
+#source("../../R/prisma.R")
+knitr::include_graphics("../figures/algo.png")
+```
+
+```{r dopamine}
+knitr::kable(dopam.anareg[,c(-3,-5)])
 ```

### 2017-12-04 R/analysis.R

@@ -29,20 +29,19 @@ require(magrittr)
 data3$d_560_adren2_v5 %>% summary
 #data3[data3$d_560_adren2_v5=="yes",15:51] %>% summary
 #data3$q_142_fatal_treatment_v5%>% summary
-#data3[data3$q_142_fatal_treatment_v5=="yes",51:100] %>% summary
 
 
 
 ### Here are some results from free treatment field of our registry - these indicate refractory anahpylaxis.
-insgesamt 3 Adrenalin-injektionen durch Hausarzt
+# insgesamt 3 Adrenalin-injektionen durch Hausarzt
 
-Ephedrin 50mg, Infusionen,Adrenalin als wiederholter Bolus (100 mikrogramm)ok, dann Adrenalinperfusor, Noradrenalinperfusor, Trendelenburglagerung
+# Ephedrin 50mg, Infusionen,Adrenalin als wiederholter Bolus (100 mikrogramm)ok, dann Adrenalinperfusor, Noradrenalinperfusor, Trendelenburglagerung
 
-Ephedrin 40mg, Magnesium 2 Ampullen, Defibrillation 3x 200Joule, Amiodaron 1 Ampulle
+# Ephedrin 40mg, Magnesium 2 Ampullen, Defibrillation 3x 200Joule, Amiodaron 1 Ampulle
 
-Atropin mehrfach 0,5 mg
-Atropin Boli, 2 minütige mechanische und medikamentöse Reanimation
-Reanimation, weitere Medi unbekannt
+# Atropin mehrfach 0,5 mg
+# Atropin Boli, 2 minütige mechanische und medikamentöse Reanimation
+# Reanimation, weitere Medi unbekannt
 
 # We had two glucagon treated anaphylaxises here !!!
 data3$q_522_glucagon_v5 %>% summary()
@@ -51,6 +50,10 @@ data3[which(data3$q_522_glucagon_v5=="yes"),200:300]
 
 # Our registry lists cases with dopamine:
 dop <- (data3$q_522_dopamine == "yes") %>% which()
-names(data3)
-data3[dop,c(11,12,66,74,366,297,43,44,45,46,47,48,49)]
+dopam.anareg <- data3[dop,c(11,12,66,74,366,
+                            241,57)]
+##### Source the files ####
+# source("R/data_handling.R")
+# source("R/prisma.R")
+# source("R/refracrory_graph.R")
 

### 2017-12-04 R/data_handling.R

@@ -1,6 +1,34 @@
 ######## Load the data ########
 # d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
 d <- read.csv("~/Documents/refractoryANA/analysis/data/raw_data/refractoryAnaphylaxis.csv")
+ d[,c(1,2,3)]
+d[,3] <- as.character(d[,3])
+
+# Clean the authors references ####
+d[c(1,8,9,10,11,12,16,17,20:43),3] <- c("[@DaSilva2017]",rep("[@Oliveira2003]",3),
+                            rep("[@DelDuca2009]",2),
+                            rep("[@Baumann2009]",2),
+                            "[@Higgins1999]",
+                            "[@Allen2000]",
+                            "[@Zaloga1986]",
+                            "[@Zweizig1994]",
+                            "[@Laxenaire1984]",
+                            "[@Bickell1984]",
+                            "[@Javeed1996]",
+                            "[@Stocche2004]",
+                            "[@Liu2005]",
+                            "[@Goddet2006]",
+                            rep("[@Schummer2008]",6),
+                            "[@Schummer2004]",
+                            "[@Liang2008]",
+                            "[@AliciaWeissgerber2007]",
+                            "[@Hussain2008]",
+                            "[@Lango2009]",
+                            "[@Raft2012]",
+                            "[@Wang2016]",
+                            "[@Gibbs2003]")
+d$Source<- gsub(d$Source,pattern = "\\[",replacement = "")
+d$Source<- gsub(d$Source,pattern = "\\]",replacement = "")
 
 ######## prepearing the data frame ######
 mean(d$age,na.rm = T)
@@ -173,9 +201,10 @@ fat <- data.frame(d[which(d$Death=="yes"),c(4,6,7,29,31,32)],
            "Fluids [L]" = c(1.5,2.5,0,2),
            "Vasopressors" = c("+","+","-","+"))
 
-pmf <- function(x){
-  ifelse(x==T,"+","-")
-}
+source("R/pmf.R")
 fat$t_steroids %<>% pmf()
 fat$t_antihistamines %<>%  pmf()
 
+#### Vasopresors analysis ####
+temp.v <- d$What.helped_gr == "vasopressors"
+d[temp.v,]

### 2017-12-04 R/pmf.R

@@ -1,0 +1,3 @@
+pmf <- function(x){
+  ifelse(x==T,"+","-")
+}

### 2017-12-04 R/prisma.R

@@ -24,7 +24,7 @@ create_graph() %>%
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 6\nnot a case report = 4\nnot anaphylaxis = 3\nanimal study = 3\njapanese language = 1",
+  add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 6\nnot a case report = 4\nnot anaphylaxis = 3\nanimal study = 3\nJapanese language = 1",
            shape="rectangle",
            width=2,
            height=1) %>%

### 2017-12-04 R/refracrory_graph.R

@@ -3,48 +3,48 @@ require(DiagrammeR)
 
 create_graph() %>%
   add_node(x = 0, y = 0,
-           label="Anaphylaxis suspected:\n1. Assess ABC\n2. Inject 300 µg adrenalin i.m.\n3. Call for help",
-           shape="rectangle",
-           width=2,
-           height=0.6) %>%
-  add_node(x = 0, y = -1,
-           label="1. Venous access\n2. 100% oxygen\n3. Crystalloids i.v\n4. Second dose of adrenalin\n5. H1+H2 Antihistamines i.v.\n6. 125 mg Prednisone i.v.",
+           label="Anaphylaxis suspected:\n1. Assess ABC\n2. Inject adrenaline i.m.",
            shape="rectangle",
            width=2.2,
-           height=1) %>%
-  add_node(x = -2, y = -2,
+           height=0.6) %>%
+  add_node(x = 0, y = -1,
+           label="No response:\n1. i.v. line and volume replacement\n2. 100% oxygen\n3. Second dose of adrenaline\n4. H1+H2 Antihistamines i.v.\n5. Corticosteroids i.v.",
+           shape="rectangle",
+           width=3,
+           height=1.1) %>%
+  add_node(x = -2.5, y = -2,
            label="Concomitant beta-blockers?",
            shape="oval",
            width=2.5) %>%
   add_node(x = 0, y = -2, label="Refractory bronchospasm?",
            shape="oval",
-           width=2) %>%
-  add_node(x = 2, y = -2, label="Refractory hypotension?",
+           width=2.2) %>%
+  add_node(x = 2.5, y = -2, label="Refractory hypotension?",
            shape="oval",
-           width=2) %>%
-  add_node(x = 4, y = -1, label="1. Observation (biphasic reaction)\n2. Adrenalin auto-injector\n3. Allergology consult",
+           width=2.5) %>%
+  add_node(x = 4, y = -0.4, label="Good clinical response:\n1. Observation (biphasic reaction)\n2. Adrenalin auto-injector\n3. Allergology consult",
+           shape="rectangle",
+           width=2.5,
+           height=0.8) %>%
+  add_node(x = -2.5, y = -3, label="consider glucagon infusion",
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = -2, y = -3, label="consider glucagon infusion",
+  add_node(x = 2.5, y = -3, label="consider vasopressors\n(norepinephrine, vasopressine)",
+           shape="rectangle",
[…]

### 2017-12-04 R/which.sources.R

@@ -1,0 +1,14 @@
+which.sources <- function(sources,pattern,variables){
+  su <- NULL
+  if(class(variables)=="list"){
+    su <- lapply(variables,function(x){
+      sources[grep(x = x, pattern=pattern)]
+    }) %>% unlist %>% unique()
+  } else {
+    s <- sources[grep(x = variables, pattern = pattern)]
+    su <- unique(s)
+  }
+
+  paste(su,collapse="; ")
+}
+

### 2017-12-06 analysis/paper/paper.Rmd
ECLS section

@@
 Patients should be transported to a critical care unit.
-The AAAAI [@Lieberman] is  suggesting a similar treatment but include dopamine and vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension.
+The AAAAI [@Lieberman] is  suggesting a similar treatment but include dopamine, vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension.
 Also the EAACI [@Muraro2014] guidelines suggest to use glucagon in case of concomitant beta-blockade.
@@
 described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia.
-The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone [@Seedat2014].
+The refractory shock was not responding to repeated adrenalin (total 2 mg), antihistaminics and hydrocortisone [@Seedat2014].
 Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred.
@@
 The European anaphylaxis register lists 18 cases where dopamine was used.
-None of these reactions were fatal and  Ring and Messmer.
+None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV.
 `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure.
@@
 
-Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Patients
+Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic-receptors.
 
-Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
+Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
+Both patients recieiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension.
+They both reacted to contrast medium and after minutes they have devloped severe anaphylaxis not responding to repeated doses of adrenaline.
+Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus releaved the refractory shock in both cases and it also sucessfully treated a recurrent biphasic reaction [@Javeed1996].
 
-Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis. [@Nag2017]*
+Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
+Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is surpressed by the beta-blockade.
+Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon what, they discuss, would be a good therapeutic attempt in their case [@Laxenaire1984].
 
 # Stopping the elicitor exposure
-In cases where the elicitor of anaphylaxis is known, and first-line  therapy fails it is advisable to stop the exposition to the likely allergen.
+In cases where the elicitor of anaphylaxis is known, and first-line therapy fails, it is advisable to stop the exposition to the likely allergen.
 Three case reports described the rapid improvement of a refractory anaphylaxis after preventing further contact with the allergen.
@@
 
-# ECLS
+# Extracorporeal life support
 
-`r (d$What.helped_gr=="ECLS") %>% which()`
-*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*
+In the most extreme cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort extracorporeal life support (ECLS) options.
+There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`].
+Two patients reacted to chlorhexidine, one to recuronium and in one case the elicitor could not be established.
+Only one of these patients was taking a beta blocker [@Yee2016].
+All these patients recieved multiple doses of adrenaline 1 -- 15 mg, vasopressors, high volume replacement therapy, coricostreroids and antihistaminics.
+Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on extracorporeal membrane oxygenation (ECMO) devices.
+After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequele.
 
-*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*
-
-*Patient with refractory anaphylaxis to Chlorhexidine coated CVC.
-10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - the return of spontaneous circulation.
-[@Wang2016] This case shows how important it is to get rid of the allergen in refractory anaphylaxis.
-Treatment with ECMO.*
+`r d[(d$What.helped_gr=="ECLS") %>% which(),]`
+life-threatening anaphylactic reactions occur in approximately 1 in 6000 anesthetic procedures and are associated with 5% mortality [@Weiss2015].
+ECMO is widely used for the treatment of cardiogenic shock [@Sidebotham2010] and may also be used in cases of anaphylaxis.
+It is especially helpful where there is refractory anaphylaxis with cardiac or respiratory arrest.
+Its use may gain time for to get rid of the elicitor (metabolize the drug or find the elicitor).
+One of the reported cases identified the clorhexidin coated cathether to be responsible for the refractory shock, and after its removal patient responded immidiately to therapy.
+The use of ECMO facilitated the elicitor discovery [@Wang2016].
 
@@
 Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously.
-This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the ….clotting …system??
-check.
+This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].
 In clinical practice it is necessary to identify patients of risk.
@@
 
+```{r wholeTab}
+knitr::kable(whole.tab)
+```
+
 ```{r dopamine}

### 2017-12-06 R/data_handling.R

@@ -34,12 +34,12 @@ d$Source<- gsub(d$Source,pattern = "\\]",replacement = "")
 mean(d$age,na.rm = T)
 d$Elicitor_gr <- rep(NA,length(d$Elicitor))
 d$Elicitor_gr[c(which(grepl("cont",d$Elicitor)==T),24)] <- "contrast"
-d$Elicitor_gr[c(4,7,13,19,20,21,38,39)] <- "unknown"
-d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,23,27,28,29,30,31,32,33,34,35,36,40,41,43)] <- "drug"
+d$Elicitor_gr[c(4,7,13,19,20,38,39)] <- "unknown"
+d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,21,23,27,28,29,30,31,32,33,34,35,36,40,41,43)] <- "drug"
 d$Elicitor_gr[c(1,3,5,15,25,37,42)] <- "other"
 
 # d$Elicitor[21] <- "platinum"
-d$Elicitor[22] <- "contrast"
+#d$Elicitor[22] <- "contrast"
 
 d$Time.after.exposure
 d$Time <- c(5,5,10,NA,5,NA,5,NA,NA,NA,NA,NA,NA,NA,15,NA,NA,5,5,NA,NA,5,5,0.2,5,5,0.5,NA,5,20,5,5,10,5,5,5,15,NA,NA,0.5,0.5,5,NA)
@@ -208,3 +208,13 @@ fat$t_antihistamines %<>%  pmf()
 #### Vasopresors analysis ####
 temp.v <- d$What.helped_gr == "vasopressors"
 d[temp.v,]
+
+#### The whole sources table #####
+
+whole.tab <- data.frame(Author =paste0(d$Author," [",d$Source,"]"),
+           Elicitor = d$Elicitor_gr,
+           Age = d$age,
+           Sex =d$sex,
+           Setting = d$Situation_gr,
+           `What helped` =d$What.helped_gr,
+           Vasopressors= d$t_vasopression)

### 2017-12-06 analysis/paper/paper.Rmd
Conclusions - preliminary

@@
 `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure.
-Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
+Adrenaline was not used in `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
 
@@
 
-`r d[(d$What.helped_gr=="ECLS") %>% which(),]`
-life-threatening anaphylactic reactions occur in approximately 1 in 6000 anesthetic procedures and are associated with 5% mortality [@Weiss2015].
+Life-threatening anaphylactic reactions occur in approximately 1 in 6000 anesthetic procedures and are associated with 5% mortality [@Weiss2015].
 ECMO is widely used for the treatment of cardiogenic shock [@Sidebotham2010] and may also be used in cases of anaphylaxis.
@@
 
+We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic aproaches.
+The cases were mostly elicited by drugs and happend during a medical procedure.
+Patients were in all cases treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v.
+and fluid replacement therapy.
+
+We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: use of mythylene blue, vasopressors, glucagon and extracorporeal life support.
+We suggest a managment algorithm for refractory anaphylaxis (see Fig.
+\@ref(fig:algo).
+
 # Expert Commentary: 500-1000 words (included in overall word count).
-<!-- **To distinguish the articles published in the Expert Review series, authors must provide an additional section entitled ‘Expert Commentary’.
-This section affords authors the opportunity to provide their interpretation of the data presented in the article and discuss the developments that are likely to be important in the future, and the avenues of research likely to become exciting as further studies yield more detailed results.
-The intention is to go beyond a conclusion and should not simply summarise the paper.
-Authors should answer the following:** -->
 
 Severe anaphylaxis is a critical medical condition requiring an immediate intervention.
-According to the international guidelines adrenaline intramuscular given is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
-However, in a few patients this intervention might not be sufficient and besides repeating the doses other therapeutical measures will need to be applied.
+According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
+However, in a few patients this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied.
 Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously.
 This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].
+
 In clinical practice it is necessary to identify patients of risk.
 Such risk factors should particulary be assessed in patients undergoing medical diagnostic and therapeutic measures.
-The measurement of tryptase enables to identify patients how are at risk for severe anaphylaxis.
+The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis.
 
@@
 # Key issues
+
 <!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->
 
-<!-- * Matrix-Assisted Laser Desorption/Ionization  Mass Spectrometry Imaging (MALDI-MSI) is able to provide a molecular insight of the samples, detecting the presence of a great variety of analytes directly on-tissue and showing their spatial distribution across the tissue section.
--->
-
-*?
-Epinephrine in appropriate doses is safe, and there are no abso- lute contraindications for its use in treating anaphylaxis.
+* Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications.
+for its use in treating anaphylaxis.
 ?

### 2017-12-06 R/refracrory_graph.R

@@ -46,6 +46,10 @@ create_graph() %>%
            shape="rectangle",
            width=2.5,
            height=0.6) %>%
+  add_node(x = 1, y = -4, label="No response:\nConsider: extracorporeal life support (CPB or ECMO)\n1. Gain time in dealing with life-threathening symptoms,\n2. Eliminate exposure to a possible (occluded) elicitor",
+           shape="rectangle",
+           width=8,
+           height=0.8) %>%
   add_edge(1,2) %>%
   add_edge(1,6) %>%
   add_edge(2,3) %>%
@@ -57,11 +61,15 @@ create_graph() %>%
   add_edge(4,9) %>%
   add_edge(10,11) %>%
   add_edge(2,10) %>%
+  add_edge(7,12) %>%
+  add_edge(8,12) %>%
+  add_edge(9,12) %>%
+  add_edge(11,12) %>%
   select_edges() %>%
   add_global_graph_attrs(
     attr = "splines",
     value = "ortho",
     attr_type = "graph") %>%
-  # render_graph()
-  export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
-                width = NULL, height = NULL)
+   render_graph()
+  # export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
+                # width = NULL, height = NULL)

### 2017-12-06 analysis/paper/paper.Rmd
Changes and template problem fix

@@
     fig_caption: yes
-    reference_docx: templates/template.docx
+    reference_docx: "../templates/template.docx"
 bibliography: references.bib
@@
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
 ```{r, setup, echo = FALSE,include=F}
@@
 
-#library(refractoryANA)
 require(magrittr,quietly = T)
-# source("../../R/data_handling.R")
 devtools::load_all(".",quiet = T)
@@
 
-## Epidemiology
-
 The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
 It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015].
-Anaphylaxis in children is in the majority of cases food dependent and most likely more frequently less severe than episodes in adulthood (Ref).
+Anaphylaxis in children is in the majority of cases food dependent and most likely less severe than episodes in adulthood [@Braganza2005].
 Cases of refractory anaphylaxis are extremely rare.
-The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3-5% of all cases of anaphylaxis.
-
-## Risk factors for anaphylaxis
+The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3--5% of all cases of anaphylaxis.
 
@@
 
-## Management
-
 The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.
-Until to day the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
-However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are for ethical reasons limited.
+To date, the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
+However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are limited for ethical reasons.
 Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.
@@
 In the present analysis we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
-
-<!-- We compared the successful procedures from the case reports to ??? -->
-
-<!-- ## Eligibility criteria -->
-<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale.
--->
-
-<!-- ## Information sources -->
-<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->
-
-## Literature search strategy
 
@@
 or i.m.) adrenaline.
-Study outcomes were defined as time to elicit a response and mortality.
 Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated.
@@
 The review protocol was not established prior to the review and the review did not undergo preregistration.
-
-<!-- ## Data items -->
-<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->
-
-<!-- ## Risk of bias in individual studies -->
-<!-- Describe methods used for assessing the risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis.
--->
-
-<!-- ## Summary measures -->
-
-<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->
-
-<!-- ## Synthesis of results -->
-
-<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->
-
-<!-- ## Risk of bias across studies  -->
-
-<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->
-
-<!-- ## Additional analyses  -->
-
-<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->
-
-```{r demograph}
-knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1)
-```
 
@@
 The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM).
-Surprisingly, no cases of insect venom anaphylaxis were found among these reports.
-While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 – 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 – 10% of the perioperative fatalities.
+Surprisingly, no cases of insect venom anaphylaxis were found among these reports (see Table \@ref(tab:demograph)).
+While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities.
 [@Nag2017]
 
-<!-- # Firs sings in anesthesia setting of anaphylaxis -->
 An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
-[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.
+[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked with increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].
 
@@
 
-Most of the treatment strategies were according to the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed.
+Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed.
 We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
 
-Based on the analyzed reports, ambulatory patients were first treated with adrenaline were patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, followed by steroid and antihistaminic agents.
-Epinephrine was given in 100% of cases in various doses.
+Based on the analyzed reports, ambulatory patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents.
+Epinephrine was given in all cases in various doses.
 
@@
 
-4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.
-
-```{r tabFatal}
-rownames(fat) <- NULL
-colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
-knitr::kable(fat)
-```
+4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table \@ref(tab:tabFatal)
 
@@
 
-Clinicians must be aware that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].
+Clinicians must be aware that methylene blue may provoke false pulse-oximetry readings as the blue dye mimics cyanosis [@Kessler1986].
 
@@
 In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated.
-
 Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases.
@@
 [`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
-[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
+[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`r which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
 Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug.
-One patient rected to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].
+One patient reacted to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].
 
@@
 Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
-Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].
+Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].
 
@@
 
-One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
+One case report described the use of military anti-shock trousers to perform external vasopression on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
 Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.
 
-The European anaphylaxis register lists 18 cases where dopamine was used.
+The European anaphylaxis register lists 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`).
 None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV.
 `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure.
-Adrenaline was not used in `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
+Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
 
@@
 
-Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic-receptors.
-
+Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
 Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
@@
 Only one of these patients was taking a beta blocker [@Yee2016].
-All these patients recieved multiple doses of adrenaline 1 -- 15 mg, vasopressors, high volume replacement therapy, coricostreroids and antihistaminics.
-Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on extracorporeal membrane oxygenation (ECMO) devices.
+All these patients recieved multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, coricostreroids and antihistaminics.
+Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device.
 After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequele.
@@
 It is especially helpful where there is refractory anaphylaxis with cardiac or respiratory arrest.
-Its use may gain time for to get rid of the elicitor (metabolize the drug or find the elicitor).
+Its use may gain time for to eliminate the responsible elicitor (metabolize the drug or find the occluded elicitor).
 One of the reported cases identified the clorhexidin coated cathether to be responsible for the refractory shock, and after its removal patient responded immidiately to therapy.
@@
 # Conclusion
-<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article.
-Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->
 
-We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic aproaches.
-The cases were mostly elicited by drugs and happend during a medical procedure.
-Patients were in all cases treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v.
+We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches.
+The cases were most frequently elicited by drugs and happend during a medical procedure.
+All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v.
 and fluid replacement therapy.
 
-We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: use of mythylene blue, vasopressors, glucagon and extracorporeal life support.
+We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support.
 We suggest a managment algorithm for refractory anaphylaxis (see Fig.
-\@ref(fig:algo).
+\@ref(fig:algo)) that is supplementary to the established protocols.
 
-# Expert Commentary: 500-1000 words (included in overall word count).
+# Expert Commentary
 
@@
 However, in a few patients this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied.
-Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously.
+Such cases are regarded as refractory anaphylaxis, which are particularly seen, when the elicitor of the reaction is reaching the organism intravenously.
 This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].
@@
 
-* Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications.
-for its use in treating anaphylaxis.
-?
-Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.*[@Fineman2015]
-* Glucagon
-* Vasopressors
-* ECLS
+* Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications and should be given as soon as anaphylaxis is suspected. 
+* Glucagon infusion should be considered when patients recieve beta-blockers.
+* Methylene blue and Vasopressors should be considered together with volume replacement therapy in cases of refractory hypotension. 
 *
 
-# References
+# Annotated bibliography
 <div id="refs"></div>
@@
 
-## Annotated bibliography
 <!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->
@@
 # Figures and Tables
-<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->
 
 ```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
-#source("../../R/prisma.R")
 knitr::include_graphics("../figures/prism_flow.png")
@@
 ```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
-#source("../../R/prisma.R")
 knitr::include_graphics("../figures/algo.png")
@@
 
-```{r wholeTab}
-knitr::kable(whole.tab)
+```{r demograph}
+knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1,
+             caption = "Demography of refractory anaphylaxis cases.")
 ```
 
-```{r dopamine}
-knitr::kable(dopam.anareg[,c(-3,-5)])
+```{r tabFatal}
+rownames(fat) <- NULL
+colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
+knitr::kable(fat,
+             caption = "Fatal anaphylaxis cases, not responsive to any form of therapy.")
 ```
 
-<!-- ### Colophon -->
-
-<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-# devtools::session_info()
+```{r wholeTab}
+knitr::kable(whole.tab,
+             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, effective interventions and therapy.")
 ```
-
-<!-- The current Git commit details are: -->
-
-```{r}
-# what commit is this file at? You may need to change the path value
-# if your Rmd is not in analysis/paper/
-# git2r::repository("../..")
-```

### 2017-12-06 R/analysis.R

@@ -52,8 +52,6 @@ data3[which(data3$q_522_glucagon_v5=="yes"),200:300]
 dop <- (data3$q_522_dopamine == "yes") %>% which()
 dopam.anareg <- data3[dop,c(11,12,66,74,366,
                             241,57)]
-##### Source the files ####
-# source("R/data_handling.R")
-# source("R/prisma.R")
-# source("R/refracrory_graph.R")
+
+knitr::kable(dopam.anareg[,c(-3,-5)])
 

### 2017-12-06 R/refracrory_graph.R

@@ -30,7 +30,7 @@ create_graph() %>%
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = 2.5, y = -3, label="consider vasopressors\n(norepinephrine, vasopressine)",
+  add_node(x = 2.5, y = -3, label="consider vasopressors\n(norepinephrine, vasopressin)\nand methylene blue",
            shape="rectangle",
            width=2.3,
            height=0.6) %>%
@@ -67,9 +67,13 @@ create_graph() %>%
   add_edge(11,12) %>%
   select_edges() %>%
   add_global_graph_attrs(
+    attr = "fontcolor",
+    value = "black",
+    attr_type = "graph") %>%
+  add_global_graph_attrs(
     attr = "splines",
     value = "ortho",
     attr_type = "graph") %>%
-   render_graph()
-  # export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
-                # width = NULL, height = NULL)
+   # render_graph()
+  export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
+                width = NULL, height = NULL)

### 2017-12-07 analysis/paper/paper.Rmd
Graphs, abstract, tables

@@
 abstract: |
-  (maximum 200 words).
+  **Introduction:**  Refractory anaphylaxis cases do not respond to treatment with intramuscular adrenaline.
+The incidence of refractory anaphylaxis is estimated to range between 3--5% of all cases of anaphylaxis.
+The risk factors for refractory anaphylaxis are unknown.
+In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
 
-  **Introduction:**  background; objectives; data sources; study eligibility criteria, participants, and interventions; study appraisal and synthesis methods; results; limitations; conclusions and implications of key findings.
-Authors are required to describe the significance of the topic under discussion.
-Areas covered - Authors are required to describe the research discussed and the literature search undertaken.
-
-  **Expert commentary:** - The author's expert view on the current status of the field under discussion. References must not be included in the abstract.
+  **Expert commentary:** According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment but may be insufficient.
+Therefore defined standardized treatment protocols for such cases of refractory anaphylaxis are needed to optimize the treatment.
+Point-of-care diagnostics may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
 keywords: |
@@
 
-Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+Department of Dermatology, Venerology, and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
 
@@
 # Introduction
-Anaphylaxis is a potentially life-threatening mast cell dependent reaction to  most frequently food, drugs, insect venom.
+Anaphylaxis is a potentially life-threatening mast cell-dependent reaction to most frequently food, drugs, insect venom.
 Anaphylaxis is a clinical diagnosis.
-It is highly likely if there is an acute onset of skin symptoms in conjuction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
+It is highly likely if there is an acute onset of skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
 The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast the term refractory anaphylaxis has not established.
+Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast, the term refractory anaphylaxis has not established.
 For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
@@
 
-In the present analysis we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
+In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
 
@@
 This information was available in all cases.
-All of the studies were case reports therefore the evidence level was classified as low.
+All of the studies were case reports, therefore, the evidence level was classified as low.
 The review protocol was not established prior to the review and the review did not undergo preregistration.
@@
 A major risk factor for refractory anaphylaxis is perioperative anaphylaxis.
-In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occured during a diagnostic or surgical procedure.
+In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occurred during a diagnostic or surgical procedure.
 The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM).
@@
 
-Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed.
+Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed (Table \@ref(tab:wholeTab)).
 We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
@@
 
-<!-- In Table X we provided therapeutic approaches used in the treatment of anaphylaxis.  -->
-
-4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table \@ref(tab:tabFatal)
+4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table \@ref(tab:tabFatal).
 
@@
 The AAAAI [@Lieberman] is suggesting a similar treatment but include dopamine, vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension.
-Also the EAACI [@Muraro2014] guidelines suggest to use glucagon in case of concomitant beta-blockade.
+Also, the EAACI [@Muraro2014] guidelines suggest using glucagon in case of concomitant beta-blockade.
 
@@
 
-The use of methylene blue in current management of anaphylaxis is scarce.
+The use of methylene blue in the current management of anaphylaxis is scarce.
 We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.
@@
 [`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`r which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
-Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug.
-One patient reacted to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].
+Drugs given perioperatively were the eliciting factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug.
+One patient reacted to systemic immune therapy (SIT) injection that was inadvertently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].
 
@@
 The refractory shock was not responding to repeated adrenalin (total 2 mg), antihistaminics and hydrocortisone [@Seedat2014].
-Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred.
+Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when the biphasic reaction occurred.
 Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014].
@@
 
-One case report described the use of military anti-shock trousers to perform external vasopression on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
+One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
 Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.
@@
 
-Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
+Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
 Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
-Both patients recieiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension.
-They both reacted to contrast medium and after minutes they have devloped severe anaphylaxis not responding to repeated doses of adrenaline.
-Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus releaved the refractory shock in both cases and it also sucessfully treated a recurrent biphasic reaction [@Javeed1996].
+Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension.
+They both reacted to contrast medium and after minutes they have developed severe anaphylaxis not responding to repeated doses of adrenaline.
+Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].
 
 Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
-Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is surpressed by the beta-blockade.
+Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade.
 Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon what, they discuss, would be a good therapeutic attempt in their case [@Laxenaire1984].
@@
 There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`].
-Two patients reacted to chlorhexidine, one to recuronium and in one case the elicitor could not be established.
+Two patients reacted to chlorhexidine, one to rocuronium and in one case the elicitor could not be established.
 Only one of these patients was taking a beta blocker [@Yee2016].
-All these patients recieved multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, coricostreroids and antihistaminics.
-Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device.
-After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequele.
+All these patients received multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, corticosteroids, and antihistaminics.
+Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device.
+After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequelae.
 
@@
 It is especially helpful where there is refractory anaphylaxis with cardiac or respiratory arrest.
-Its use may gain time for to eliminate the responsible elicitor (metabolize the drug or find the occluded elicitor).
-One of the reported cases identified the clorhexidin coated cathether to be responsible for the refractory shock, and after its removal patient responded immidiately to therapy.
+Its use may gain time to eliminate the responsible elicitor (metabolize the drug or find the occluded elicitor).
+One of the reported cases identified the chlorhexidine coated catheter to be responsible for the refractory shock, and after its removal patient responded immediately to therapy.
 The use of ECMO facilitated the elicitor discovery [@Wang2016].
@@
 
-We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches.
-The cases were most frequently elicited by drugs and happend during a medical procedure.
+We analyzed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches.
+The cases were most frequently elicited by drugs and happened during a medical procedure.
 All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v.
@@
 
-We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support.
-We suggest a managment algorithm for refractory anaphylaxis (see Fig.
+We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first-line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support.
+We suggest a management algorithm for refractory anaphylaxis (see Fig.
 \@ref(fig:algo)) that is supplementary to the established protocols.
@@
 According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
-However, in a few patients this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied.
-Such cases are regarded as refractory anaphylaxis, which are particularly seen, when the elicitor of the reaction is reaching the organism intravenously.
-This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].
+However, in a few patients, this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied.
+Such cases are regarded as refractory anaphylaxis and are particularly seen when the elicitor of the reaction is reaching the organism intravenously.
+This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancement of vascular permeability [@Imamura1996].
 
-In clinical practice it is necessary to identify patients of risk.
-Such risk factors should particulary be assessed in patients undergoing medical diagnostic and therapeutic measures.
+In clinical practice, it is necessary to identify patients at risk.
+Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures.
 The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis.
 
-For the future, biomarker measurements in a bed side setting may enable doctors to identify  patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
-Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can safe life in a given patient.
+For the future, biomarker measurements at a bedside setting may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
+Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can save the life in a given patient.
 
@@
 
-<!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->
-
+* Drug elicited reactions in perioperative setting and concomitant beta-blocker use are the most important risk factors for refractory anaphylaxis.
 * Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications and should be given as soon as anaphylaxis is suspected. 
-* Glucagon infusion should be considered when patients recieve beta-blockers.
-* Methylene blue and Vasopressors should be considered together with volume replacement therapy in cases of refractory hypotension. 
-*
+* Glucagon infusion should be considered when patients receive beta-blockers.
+* Methylene blue and vasopressors i.v. should be considered together with volume replacement therapy in cases of refractory hypotension. 
+* In cases of severe refractory anaphylaxis with prolonged resuscitation an extracorporeal life support should be evaluated.
 
@@
 <div id="refs"></div>
-<!-- **A maximum of 100 references is suggested.
-Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography.
-Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->
-
-<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->
 
@@
 
+```{r wholeTab}
+knitr::kable(whole.tab,
+             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, and effective therapy interventions")
+```
+
 ```{r tabFatal}
@@
 ```
-
-```{r wholeTab}
-knitr::kable(whole.tab,
-             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, effective interventions and therapy.")
-```

### 2017-12-07 R/data_handling.R

@@ -29,11 +29,22 @@ d[c(1,8,9,10,11,12,16,17,20:43),3] <- c("[@DaSilva2017]",rep("[@Oliveira2003]",3
                             "[@Gibbs2003]")
 d$Source<- gsub(d$Source,pattern = "\\[",replacement = "")
 d$Source<- gsub(d$Source,pattern = "\\]",replacement = "")
-
+d$Author <- gsub(d$Author,pattern = " et .*",replacement = "")
+d$Author[c(18,19,21,22,28,29)] <- c(rep("Heytman",2),
+                                    "Allen","Zaloga","Liu","Goddet")
+#### Clear elicitors #####
+d$Elicitor <- as.character(d$Elicitor)
+d$Elicitor <- gsub(" .*","",d$Elicitor)
+d$Elicitor[c(1,6,7,
+             15,19,20,
+             24,25,42)] <- c("latex","bendamustine","unknown",
+                                 "menses","unknown","unknown",
+                                 "contrast","SIT","chlorhexidine"
+                           )
 ######## prepearing the data frame ######
 mean(d$age,na.rm = T)
 d$Elicitor_gr <- rep(NA,length(d$Elicitor))
-d$Elicitor_gr[c(which(grepl("cont",d$Elicitor)==T),24)] <- "contrast"
+d$Elicitor_gr[c(which(grepl("cont",d$Elicitor)==T))] <- "contrast"
 d$Elicitor_gr[c(4,7,13,19,20,38,39)] <- "unknown"
 d$Elicitor_gr[c(2,6,11,12,13,14,16,17,18,21,23,27,28,29,30,31,32,33,34,35,36,40,41,43)] <- "drug"
 d$Elicitor_gr[c(1,3,5,15,25,37,42)] <- "other"
@@ -109,8 +120,10 @@ demogr.tab <- data.frame(Age=d$age %>% split(d$Elicitor_gr) %>%
              lapply(function(x){
                which(x=="OP")  %>% length()/length(x)*100
              }) %>% unlist
-) %>% data.frame(Elicitor=rownames(.),
-                 n=d$Elicitor_gr %>% factor %>% summary() %>% unlist,.)
+           ) %>%
+             data.frame(Elicitor=rownames(.),
+                 n=d$Elicitor_gr %>% factor %>% summary() %>% unlist,
+                 .)
 
 rownames(demogr.tab) <- NULL
 
@@ -212,9 +225,9 @@ d[temp.v,]
 #### The whole sources table #####
 
[…]

### 2017-12-07 R/prisma.R

@@ -27,7 +27,7 @@ create_graph() %>%
   add_node(x = 3, y = 1, label="Full-text articles excluded\nnon-refractory = 6\nnot a case report = 4\nnot anaphylaxis = 3\nanimal study = 3\nJapanese language = 1",
            shape="rectangle",
            width=2,
-           height=1) %>%
+           height=1.1) %>%
   add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 29)",
            shape="rectangle",
            width=2,
@@ -44,6 +44,9 @@ create_graph() %>%
     attr = "splines",
     value = "spline",
     attr_type = "graph") %>%
+  add_global_graph_attrs(value="black",
+                         attr = "fontcolor",
+                         attr_type = "node") %>%
   # render_graph() %>%
   export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/prism_flow.png", file_type = "png", title = NULL,
                width = NULL, height = NULL)

### 2017-12-07 R/refracrory_graph.R

@@ -1,7 +1,7 @@
 # Refractory ANA flow diagdramm
 require(DiagrammeR)
 
-create_graph() %>%
+g <- create_graph() %>%
   add_node(x = 0, y = 0,
            label="Anaphylaxis suspected:\n1. Assess ABC\n2. Inject adrenaline i.m.",
            shape="rectangle",
@@ -66,14 +66,12 @@ create_graph() %>%
   add_edge(9,12) %>%
   add_edge(11,12) %>%
   select_edges() %>%
-  add_global_graph_attrs(
-    attr = "fontcolor",
-    value = "black",
-    attr_type = "graph") %>%
-  add_global_graph_attrs(
-    attr = "splines",
-    value = "ortho",
-    attr_type = "graph") %>%
-   # render_graph()
-  export_graph(file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
+  add_global_graph_attrs(value="black",
+                         attr = "fontcolor",
+                         attr_type = "node") %>%
+  add_global_graph_attrs(value="ortho",
+                         attr = "splines",
+                         attr_type = "graph")
+  render_graph(g)
+  export_graph(g,file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
                 width = NULL, height = NULL)

### 2017-12-07 analysis/paper/paper.Rmd
Biomarkers

@@
   - Margitta Worm, MD
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: "`r format(Sys.time(), '%d.%m.%Y')`"
 output: 
@@
 Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures.
-The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis.
+The diagnosis of anaphylaxis is based on unspecific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing.
+In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline.
+The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].
 
-For the future, biomarker measurements at a bedside setting may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
+For the future,  measurements [@Wittenberg2017] at a bedside setting may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment.
 Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can save the life in a given patient.

### 2017-12-15 analysis/paper/paper.Rmd
changes small
finished correction from Prof Worm

@@
 bibliography: references.bib
-csl: mee.csl
+csl: erci.csl
 abstract: |
-  **Introduction:**  Refractory anaphylaxis cases do not respond to treatment with intramuscular adrenaline.
-The incidence of refractory anaphylaxis is estimated to range between 3--5% of all cases of anaphylaxis.
+  **Introduction:** Patients experiencing anaphylaxis who do not recover after the treatment with intramuscular adrenaline are regarded to suffer from refractory anaphylaxis.
+The incidence of refractory anaphylaxis is estimated to range between 3--5% of anaphylaxis cases.
 The risk factors for refractory anaphylaxis are unknown.
-In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
 
-  **Expert commentary:** According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment but may be insufficient.
+  **Areas covered:** In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.
+
+  **Expert commentary:** According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and safe treatment but may be insufficient.
 Therefore defined standardized treatment protocols for such cases of refractory anaphylaxis are needed to optimize the treatment.
@@
 # Introduction
-Anaphylaxis is a potentially life-threatening mast cell-dependent reaction to most frequently food, drugs, insect venom.
+Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom.
 Anaphylaxis is a clinical diagnosis.
-It is highly likely if there is an acute onset of skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
+It is highly likely if there is an acute onset of mainly skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
 The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast, the term refractory anaphylaxis has not established.
-For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin.
+Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].
+By contrast, the term refractory anaphylaxis has not been established yet.
+For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline.
 Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e.
@@
 
-Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013].
-Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigorous exercise (e.g.
+Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mast cell activation syndromes) [@Worm2013].
+Severe anaphylaxis is more likely to be elicited in elderly patients, patients suffering from mastocytosis, and in temporal proximity to vigorous exercise (e.g.
 jogging) [@Worm2013].
@@
 The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy.
-To date, the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
+To date, the acute treatment of anaphylaxis is uniquely recommended in the international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
 However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are limited for ethical reasons.
@@
 Additionally, case reports referenced in the included sources were added.
-We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenaline.
+We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline.
 Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
-We finally included 29 Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
+We finally included `r unique(d$Sources) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
 The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
@@
 An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
-[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked with increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].
+[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked to increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].
 
@@
 Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed (Table \@ref(tab:wholeTab)).
-We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
+We report on the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.
 
-Based on the analyzed reports, ambulatory patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents.
+Based on the analyzed reports, out-patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents.
 Epinephrine was given in all cases in various doses.
@@
 
-World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011].
+The World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011].
 Patients should be transported to a critical care unit.
@@
 
-The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce Nitric oxide (NO).
+The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce nitric oxide (NO).
 NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000].
-This in effect leads to the opening of the calcium ion channels in smooth muscle tissues and subsequent loss of vascular muscle contractility [@Omar2015].
+This leads to the an opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].
 
-Methylene blue is a well-investigated drug used for treating patients with methemoglobinemia [@Schirmer2003].
+Methylene blue is a well-investigated drug to treat patients with methemoglobinemia [@Schirmer2003].
 Buzato et al.
-investigated its properties as an inhibitor of nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue was mediated by the sGC inhibition.
+investigated its properties as an inhibitor of the nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue is mediated by the sGC inhibition.
 Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].
@@
 Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013].
-Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) production which is a potent mediator of anaphylaxis [@Moritoki1992].
+The authors suggested that the inhibition of GC may result in a decrease of histamine and platelet-activating factor (PAF) production which are potent mediators of anaphylaxis [@Moritoki1992].
 
-The use of methylene blue in the current management of anaphylaxis is scarce.
+Data about the use of methylene blue in the current management of anaphylaxis are scarce.
 We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.
 
-We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension using 100 mg bolus i.v.
-and follow it with an infusion of 50 µg/kg/min.
+In case of a refractory anaphylaxis with severe persisting hypotension we propose to apply a 100 mg bolus i.v.
+followed by an infusion of 50 µg/kg/min.
 As a vector, a 100 mL of 5% dextrose can be used.
@@
 As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
-It may be considered a safe option for the treatment of anaphylaxis.
-No side effects were reported due to methylene blue use in the above cases.
+It may be considered a safe option for the treatment of anaphylaxis upon intensive care conditions.
+No side effects were reported due to methylene blue use in the reported-cases.
 
@@
 
-In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated.
+In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be considered.
 Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases.
 Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases.
-In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes.
+In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) the injection of vasopressin reversed the refractory hypotension within minutes.
 Other vasopressors included metaraminol 
@@
 
-Use of vasopressin in refractory anaphylaxis was only suggested in management recommendations from the AAAAI [@Lieberman] and was never vastly investigated.
+The use of vasopressin in refractory anaphylaxis was only suggested in the management recommendations from the AAAAI [@Lieberman] and was never vastly investigated.
 Dopamine, however, is recommended by 2 out of three guidelines.
-Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015].
+Nevertheless, dopamine might be related to higher a risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015].
 Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
@@
 described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia.
-The refractory shock was not responding to repeated adrenalin (total 2 mg), antihistaminics and hydrocortisone [@Seedat2014].
-Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when the biphasic reaction occurred.
-Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014].
+The refractory shock was not responding to repeated adrenaline (total 2 mg), antihistamines and hydrocortisone [@Seedat2014].
+A rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when the biphasic reaction occurred.
+The authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014].
 We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.
@@
 One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
-Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.
+Although this can be hardly available in critical care centers we want to point out this minimally invasive method as a possible alternative measure in achieving normotension.
 
-The European anaphylaxis register lists 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`).
+The European anaphylaxis register includes 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`).
 None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV.
@@
 
-Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
-Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
-Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension.
-They both reacted to contrast medium and after minutes they have developed severe anaphylaxis not responding to repeated doses of adrenaline.
-Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].
+Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lung express beta-adrenergic receptors.
+Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
+Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomittant use of atenolol for arterial hypertension.
+Both patients reacted to contrast medium and after minutes developed severe anaphylaxis not responding to repeated doses of adrenaline.
+An infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].
 
-Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.
-Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade.
-Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon what, they discuss, would be a good therapeutic attempt in their case [@Laxenaire1984].
+Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014] Therefore, it is especially useful in patients who are on long term beta-blocker treatments.
+Patients with concomittant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade.
+Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon but discuss this therapeutic option positively as well [@Laxenaire1984].
 
@@
 A 15-year-old boy who was exposed to cow's milk underwent near-fatal anaphylaxis with refractory hypotension.
-After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylaxis symptoms and patient recovered fully.
-Authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].
+After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylactic symptoms and the patient recovered completely.
+These authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].
 
@@
 
-In the most extreme cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort extracorporeal life support (ECLS) options.
+In the most severe cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort of extracorporeal life support (ECLS) options.
 There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`].
@@
 Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device.
-After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequelae.
+After stabilizing of their condition, patients were weaned off from these supportive measures without neurological sequelae.
 
@@
 
-We analyzed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches.
-The cases were most frequently elicited by drugs and happened during a medical procedure.
+We analyzed case reports from literature and own data from our anaphylaxis registry of refractory anaphylaxis and critically evaluated the undertaken therapeutic approaches.
+These cases of refractory anaphylaxis were most frequently elicited by drugs and happened during a medical procedure.
 All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v.

### 2018-02-09 analysis/paper/paper.Rmd
changed the template

@@
 ---
-title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
+title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a review of case reports'
 author:
@@
     fig_caption: yes
-    reference_docx: "../templates/template.docx"
+    reference_docx: "../templates/templateNumberedSections.docx"
 bibliography: references.bib
@@
 
+**Keywords:** anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock
+
 # Introduction
 Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom.
-Anaphylaxis is a clinical diagnosis.
-It is highly likely if there is an acute onset of mainly skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006].
+Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) Acute onset of an illness with generalized skin / mucosal involvemt with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006].
 The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].
 By contrast, the term refractory anaphylaxis has not been established yet.
@@
 Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
-We finally included `r unique(d$Sources) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
+We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
 The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
@@
 The review protocol was not established prior to the review and the review did not undergo preregistration.
+It is important to mention that as a review of case reports this study did not use control groups to compare outcomes and its statistical validity is therefore limited.
 
-# Risk factors for refractory anaphylaxis
+## Risk factors for refractory anaphylaxis
 
@@
 
-# Therapy of refractory anaphylaxis
+## Therapy of refractory anaphylaxis
 
@@
 
-# Methylene blue for the treatment of refractory hypotension
+## Methylene blue for the treatment of refractory hypotension
 
@@
 NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000].
-This leads to the an opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].
+This leads to the opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].
 
@@
 
-# Vasopressors
+## Vasopressors
 
@@
 Dopamine, however, is recommended by 2 out of three guidelines.
-Nevertheless, dopamine might be related to higher a risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015].
+Nevertheless, dopamine might be related to a higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015].
 Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension.
@@
 
-# Glucagon in anaphylaxis with concomitant beta blockers
+## Glucagon in anaphylaxis with concomitant beta blockers
 
@@
 
-# Stopping the elicitor exposure
+## Stopping the elicitor exposure
 In cases where the elicitor of anaphylaxis is known, and first-line therapy fails, it is advisable to stop the exposition to the likely allergen.
@@
 
-# Extracorporeal life support
+## Extracorporeal life support
 
@@
 
-# Conclusion
+## Conclusion
 
@@
 Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures.
-The diagnosis of anaphylaxis is based on unspecific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing.
+The diagnosis of anaphylaxis is based on non-specific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing.
 In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline.
 The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].
+
+# Five year View
 
@@
 knitr::kable(whole.tab,
-             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, and effective therapy interventions")
+             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions")
 ```

### 2018-02-21 analysis/paper/paper.Rmd
revision 1st round

@@
 Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) Acute onset of an illness with generalized skin / mucosal involvemt with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006].
+The definition of anaphylaxis remains controverse [@Lieberman2015].
 The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].
 
-By contrast, the term refractory anaphylaxis has not been established yet.
+By contrast, the definition of refractory anaphylaxis has not been established yet.
 For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline.
@@
 rapid increase in blood pressure, bronchodilatation, tachycardia).
+Previously authors suggested using the term in cases of persistent severe shock symptoms after 10 minutes of adequate resustitation [@Gouel2016].
 
-It is important to distinguish refractory anaphylaxis from recurrent anaphylaxis.
+It is important to distinguish refractory anaphylaxis (which is unresponsive to treatment) from recurrent anaphylaxis which can be defined as multiple episodes of anaphylaxis over a period of time that responsed to treatment with adrenaline [@Mullins2003].
 In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
@@
 Cases of refractory anaphylaxis are extremely rare.
-The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3--5% of all cases of anaphylaxis.
+The incidence of refractory anaphylaxis was not investigated earlier and we estimate it to range between 3--5% of all cases of anaphylaxis (based on our clinical experience).
 
-Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mast cell activation syndromes) [@Worm2013].
+Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mast cell activation syndromes) [@Worm2018].
 Severe anaphylaxis is more likely to be elicited in elderly patients, patients suffering from mastocytosis, and in temporal proximity to vigorous exercise (e.g.
@@
 The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017.
-The search was conducted on 16.11.2017 and returned 150 results.
+The search was conducted on 16.11.2017 and returned 151 results.
 All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study.
@@
 Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
-We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis.
 The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
@@
 
+We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis (Table \@ref(tab:demo))
+
 ## Risk factors for refractory anaphylaxis
@@
 The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM).
+Indirectly concomitant treatment with protimne-containing insulins might predispose to refractory anaphylaxis as these patiens have 40 to 50 times greater risk for developing anaphylaxis perioperatively [@Stewart1984; @DrugAllergy2010].
 Surprisingly, no cases of insect venom anaphylaxis were found among these reports (see Table \@ref(tab:demograph)).
@@
 
-An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations.
-[@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked to increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].
+The patomechanisms which are responsible for refractory anaphylaxis are not well understood.
+There are however theories why extensive therapeutic efforts fail in certain cases.
+Figure \@ref(fig:pat) summarizes the potential factors contributing to refctorinness of anaphylaxis.
+These can be divided into patient dependent (e.g.
+concomitant coronary disease, older age or bronchial hyperresponsiveness) or therapy-dependent (e.g.
+suboptimal management, concomitant beta blocker use, or idopathic triggers).
+
+An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse [@Gouel2016], which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations [@Baumann2009].
+In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline  use has been linked to increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].
 
@@
 
-The European anaphylaxis register includes 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`).
+<!-- The European anaphylaxis register includes 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`).
 None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV.
 `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure.
-Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%
+Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%  -->
 
@@
 ## Conclusion 
-
-We analyzed case reports from literature and own data from our anaphylaxis registry of refractory anaphylaxis and critically evaluated the undertaken therapeutic approaches.
+We analyzed case reports from the literature and critically evaluated the undertaken therapeutic approaches.
 These cases of refractory anaphylaxis were most frequently elicited by drugs and happened during a medical procedure.
@@
 
-```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
+```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources.
+The additional 18 papers were included as a result of screening the referenced sources in the initial 151 screened papres.
+We screened the pubmed database using the search terms 'refractory anaphylaxis' with the help of the RISmed package for R.
+Only Case reports were included."}
 knitr::include_graphics("../figures/prism_flow.png")
@@
 
+```{r pato, echo=F,fig.cap="The patophyisiology of refractory anaphylaxis.
+The proposed mechanisms of anaphylaxis reaction non responding to therapy are divided into patient dependent and iatrogenic factors."}
+knitr::include_graphics("../figures/pato.png")
+```
+
+```{r demo }
+table_demo %>% cbind(sex=c("male","female"),.) %>% knitr::kable(
+  caption = "Summary of the refractory anaphylxis cases included into the analysis according to concomitant diseases (arterial hypertension and diabetes (either type I or II) expressed as a fraction of included cases).
+TTR - median time to reaction after exposure to the suspected allergen expressed in minutes.
+Age is a mean value across evaluated cases in each group.",
+                            col.names = c("Sex","n","Age","TTR [min]","Hypertension [%]","Diabetes [%]"),
+                            row.names = F,
+                            digits = 1)
+```
+
 ```{r demograph}
-knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1,
-             caption = "Demography of refractory anaphylaxis cases.")
+knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male [%]","Perioperative [%]"),digits = 1,
+             caption = "Refractory anaphylaxis cases broken down according to the elicitor of the reaction.
+Male - percentage of male cases to all cases ina griven group.
+Perioperative - fraction of anaphylaxis episode that happened either during surgery or in temporal proximity to a surgical procedure.
+The eliciting drugs were: cefuroxime, chlorhexidine, protamine, aprotinin, succinylocholine, gelofusine, recuronium, chemotherapy,  amoxicillin, ketokonazole, metamizole")
 ```
@@
 knitr::kable(whole.tab,
-             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions")
+             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions.
+Cases where the reaction was fatal and all therapeutic options fialed were indicated with the asterix (*).
+Other cases resulted with full recovery of the patient.
+")
 ```

### 2018-02-21 R/analysis.R

@@ -1,6 +1,6 @@
 # The search was performed on 16.11.2017
-require(RISmed)
 require(magrittr)
+require(RISmed)
 res <- EUtilsSummary("refractory anaphylaxis",
 type="esearch",
 db="pubmed",

### 2018-02-21 R/data_handling.R

@@ -1,11 +1,15 @@
 ######## Load the data ########
 # d <- read.csv("../data/raw_data/refractoryAnaphylaxis.csv")
-d <- read.csv("~/Documents/refractoryANA/analysis/data/raw_data/refractoryAnaphylaxis.csv")
+d <- read.csv("~/Documents/refractoryANA/analysis/data/raw_data/refractoryAnaphylaxis.csv",colClasses = "character")
  d[,c(1,2,3)]
 d[,3] <- as.character(d[,3])
 
+d2<- read.csv("analysis/data/raw_data/REfractory anaphylaxis - Arkusz1.csv",colClasses = "character")
+
+
+d <- rbind(d,d2[45:47,1:26])
 # Clean the authors references ####
-d[c(1,8,9,10,11,12,16,17,20:43),3] <- c("[@DaSilva2017]",rep("[@Oliveira2003]",3),
+d[c(1,8,9,10,11,12,16,17,20:43,44:46),3] <- c("[@DaSilva2017]",rep("[@Oliveira2003]",3),
                             rep("[@DelDuca2009]",2),
                             rep("[@Baumann2009]",2),
                             "[@Higgins1999]",
@@ -26,7 +30,8 @@ d[c(1,8,9,10,11,12,16,17,20:43),3] <- c("[@DaSilva2017]",rep("[@Oliveira2003]",3
                             "[@Lango2009]",
                             "[@Raft2012]",
                             "[@Wang2016]",
-                            "[@Gibbs2003]")
+                            "[@Gibbs2003]",
+                            rep("[@Gouel]",3))
 d$Source<- gsub(d$Source,pattern = "\\[",replacement = "")
 d$Source<- gsub(d$Source,pattern = "\\]",replacement = "")
 d$Author <- gsub(d$Author,pattern = " et .*",replacement = "")
@@ -37,15 +42,17 @@ d$Elicitor <- as.character(d$Elicitor)
 d$Elicitor <- gsub(" .*","",d$Elicitor)
 d$Elicitor[c(1,6,7,
              15,19,20,
-             24,25,42)] <- c("latex","bendamustine","unknown",
+             24,25,42,44:46)] <- c("latex","bendamustine","unknown",
[…]

### 2018-02-21 R/pato.R

@@ -1,0 +1,85 @@
+#pathofiz diagdramm
+require(DiagrammeR)
+
+create_graph() %>%
+  add_node(x = -2.8, y = -1,
+           label="Impaired right\nventricular function",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = -2.8, y = -2, label="Coronary artery disease",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = 0, y = -2, label="Failure to restore\nsufficient cardiac output",
+           shape="rectangle",
+           width=2.5) %>%
+  add_node(x = 0, y = -3, label="Insufficient tissue perfusion",
+           shape="rectangle",
+           width=2) %>%
+  add_node(x = 0, y = -4, label="Multiorgan failure",
+           shape="rectangle",
+           width=2) %>%
+  add_node(x = 3, y = -4, label="Concomitant beta-blockers",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = 3, y = -3, label="Insufficient volume\nresubstitution, vasopression",
+           shape="rectangle",
+           width=2.2,
+           height=0.8) %>%
+  add_node(x = 0, y = 0, label="Persistent allergen contact",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 0, y = -1, label="Prolonged vasodialation",
+           shape="rectangle",
+           width=2,
+           height=0.6) %>%
+  add_node(x = 3, y = 0, label="Iatrogenic factors",
+           shape="plaintext",
+           width=2,
+           height=0.6) %>%
+  add_node(x = -2.8, y = 0, label="Patient dependent\n factors",
+           shape="plaintext",
+           width=2,
+           height=0.6) %>%
+  add_node(x = -2.8, y = -3, label="Older age",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = 3, y = -2, label="Late epinephrine injection",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
+  add_node(x = 3, y = -1, label="Failure to cease\nallergen exposition",
+           shape="rectangle",
+           width=2.2,
+           height=0.6) %>%
[…]

### 2018-02-21 R/prisma.R

@@ -3,11 +3,11 @@ require(DiagrammeR)
 
 create_graph() %>%
   add_node(x = -1.2, y = 4,
-           label="Records identified through\ndatabase searching\n(n = 150)",
+           label="Records identified through\ndatabase searching\n(n = 151)",
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = 1.2, y = 4, label="Additional records idientified\nthrough references\n(n = 18)",
+  add_node(x = 1.2, y = 4, label="Additional records idientified\nthrough references from screened\nreports (n = 18)",
            shape="rectangle",
            width=2.2,
            height=0.6) %>%
@@ -28,7 +28,7 @@ create_graph() %>%
            shape="rectangle",
            width=2,
            height=1.1) %>%
-  add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 29)",
+  add_node(x = 0, y = 0, label="Studies included in\nqualitative synthesis\n(n = 30)",
            shape="rectangle",
            width=2,
            height=0.6) %>%

### 2018-02-23 analysis/paper/paper.Rmd
changes post 1st rev round + Prof Worm

@@
 
-Department of Dermatology, Venerology, and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+Division of Allergy and Immunology, Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
 
@@
 Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom.
-Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) Acute onset of an illness with generalized skin / mucosal involvemt with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006].
+Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) acute onset of an illness with generalized skin or mucosal involvement with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006].
 The definition of anaphylaxis remains controverse [@Lieberman2015].
@@
 rapid increase in blood pressure, bronchodilatation, tachycardia).
-Previously authors suggested using the term in cases of persistent severe shock symptoms after 10 minutes of adequate resustitation [@Gouel2016].
+Previously authors suggested using the term in cases of persistent severe shock after 10 minutes of adequate resustitation [@Gouel2016].
 
-It is important to distinguish refractory anaphylaxis (which is unresponsive to treatment) from recurrent anaphylaxis which can be defined as multiple episodes of anaphylaxis over a period of time that responsed to treatment with adrenaline [@Mullins2003].
+It is important to distinguish refractory anaphylaxis (which is unresponsive to treatment) from recurrent anaphylaxis which can be defined as multiple episodes of anaphylaxis over a period of time that responded to treatment with adrenaline [@Mullins2003].
 In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab.
@@
 
-The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
+The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years with an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].
 It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015].
@@
 Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included.
-The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).
+The detailed inclusion flowchart is illustrated in Fig.
+\@ref(fig:prismaflow).
 
@@
 
-We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis (Table \@ref(tab:demo))
+We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of refractory anaphylaxis (Table \@ref(tab:demo)).
 
@@
 Indirectly concomitant treatment with protimne-containing insulins might predispose to refractory anaphylaxis as these patiens have 40 to 50 times greater risk for developing anaphylaxis perioperatively [@Stewart1984; @DrugAllergy2010].
-Surprisingly, no cases of insect venom anaphylaxis were found among these reports (see Table \@ref(tab:demograph)).
+Surprisingly, no cases of insect venom anaphylaxis were found among these reports (Table \@ref(tab:demograph)).
 While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities.
@@
 
-The patomechanisms which are responsible for refractory anaphylaxis are not well understood.
+The pathomechanisms which are responsible for refractory anaphylaxis are not well understood.
 There are however theories why extensive therapeutic efforts fail in certain cases.
-Figure \@ref(fig:pat) summarizes the potential factors contributing to refctorinness of anaphylaxis.
+Fig.
+\@ref(fig:pato) summarizes the potential factors contributing to refctorinness of anaphylaxis.
 These can be divided into patient dependent (e.g.
@@
 
-As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]).
+As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discoloration, and hyperbilirubinemia [@Evora2007]).
 It may be considered a safe option for the treatment of anaphylaxis upon intensive care conditions.
-No side effects were reported due to methylene blue use in the reported-cases.
+No side effects were reported due to methylene blue use in the reported cases.
 
@@
 
-One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
+One case report described the use of military anti-shock trousers to perform external vasopression on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984].
 Although this can be hardly available in critical care centers we want to point out this minimally invasive method as a possible alternative measure in achieving normotension.
@@
 Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lung express beta-adrenergic receptors.
-Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996].
+Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension [@Zaloga1986; @Javeed1996].
 Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomittant use of atenolol for arterial hypertension.
@@
 
-Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014] Therefore, it is especially useful in patients who are on long term beta-blocker treatments.
+Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014].
+Therefore, it is especially useful in patients who are on long term beta-blocker treatments.
 Patients with concomittant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade.
@@
 After stabilizing of their condition, patients were weaned off from these supportive measures without neurological sequelae.
+Gouel-Chéron et al.
+reported a fifth case of refractory anaphylaxis where the patient died on the 5th day after introducing ECLS [@Gouel].
 
@@
 We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first-line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support.
-We suggest a management algorithm for refractory anaphylaxis (see Fig.
+We suggest a management algorithm for refractory anaphylaxis (Fig.
 \@ref(fig:algo)) that is supplementary to the established protocols.
@@
 Severe anaphylaxis is a critical medical condition requiring an immediate intervention.
-According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
+According to the international guidelines, adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient.
 However, in a few patients, this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied.
@@
 The diagnosis of anaphylaxis is based on non-specific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing.
-In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline.
-The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].
+In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-care test could confirm the suspicion in order to introduce the immediate therapy with adrenaline.
+The measurement of peak tryptase levels (with the cut off value of 12 µg/L) could be used for confirming the diagnosis of anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].
 
@@
 Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can save the life in a given patient.
+
+Aditionally, based on clinical features, we will be able to predict which patients are at risk of developing severe anaphylaxis.
+Therefore extraordinary preventive measures could be undertaken in such patients beforehand.
 
@@
 
-```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
-knitr::include_graphics("../figures/algo.png")
+```{r pato, echo=F,fig.cap="The overview of factors influencing the onset of refractory anaphylaxis. "}
+knitr::include_graphics("../figures/pato.png")
 ```
 
-```{r pato, echo=F,fig.cap="The patophyisiology of refractory anaphylaxis.
-The proposed mechanisms of anaphylaxis reaction non responding to therapy are divided into patient dependent and iatrogenic factors."}
-knitr::include_graphics("../figures/pato.png")
+```{r algo, echo=F,fig.cap="Proposed algorithm for refractory anaphylaxis management"}
+knitr::include_graphics("../figures/algo.png")
 ```
@@
              caption = "Refractory anaphylaxis cases broken down according to the elicitor of the reaction.
-Male - percentage of male cases to all cases ina griven group.
+Male - percentage of male cases to all cases in a given group.
 Perioperative - fraction of anaphylaxis episode that happened either during surgery or in temporal proximity to a surgical procedure.
-The eliciting drugs were: cefuroxime, chlorhexidine, protamine, aprotinin, succinylocholine, gelofusine, recuronium, chemotherapy,  amoxicillin, ketokonazole, metamizole")
+The eliciting drugs were: cefuroxime, chlorhexidine, protamine, aprotinin, succinylocholine, gelofusine, recuronium, chemotherapy,  amoxicillin, ketokonazole, metamizole.")
 ```
@@
 knitr::kable(whole.tab, 
-             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions.
+             caption = "List of reports included into this study, summarizing their elicitors, and effective therapy interventions.
 Cases where the reaction was fatal and all therapeutic options fialed were indicated with the asterix (*).
 Other cases resulted with full recovery of the patient.
-")
+'Therapy' column describes the approach which lead to a rapid recovery in a given case.
+HT - arterial hypertension and DM - diabetes mellitus are reported (other concomitant diseases are not included in this table).
+OP - perioperative anaphylaxis.",
+            col.names = c("Author","Elicitor","Age","Sex","Seting","Therapy","Conc. disease","Vaopressors"))
 ```

### 2018-02-23 R/data_handling.R

@@ -66,22 +66,22 @@ d$Situation_gr <- c("OP",
                     "OP",
                     "OP",
                     "ambulatory",
-                    "not hospital",
+                    "ambulatory",
                     "chemo",
                     "OP",
-                    "unknown",
-                    "unknown",
-                    "unknown",
+                    "CT",
+                    "CT",
+                    "CT",
                     "OP",
                     "OP",
                     "OP","unknown",
-                    "not hospital",
+                    "ambulatory",
                     "OP",
                     "OP",
                     "OP",
                     "OP",
                     "unknown",
-                    "unknown",
+                    "OP",
                     "CT",
                     "chemo",
                     "OP",

### 2018-02-23 R/pato.R

@@ -72,10 +72,10 @@ create_graph() %>%
   add_edge(13,9) %>%
   add_edge(14,9) %>%
   add_edge(15,4) %>%
-  select_edges() %>%
+  # select_edges() %>%
   add_global_graph_attrs(
     attr = "splines",
-    value = "spline",
+    value = "ortho",
     attr_type = "graph") %>%
   add_global_graph_attrs(value="black",
                          attr = "fontcolor",

### 2018-02-23 R/prisma.R

@@ -7,20 +7,20 @@ create_graph() %>%
            shape="rectangle",
            width=2,
            height=0.6) %>%
-  add_node(x = 1.2, y = 4, label="Additional records idientified\nthrough references from screened\nreports (n = 18)",
+  add_node(x = 1.2, y = 4, label="Additional records idientified\nthrough references from screened\nreports (n = 19)",
            shape="rectangle",
            width=2.2,
            height=0.6) %>%
-  add_node(x = 0, y = 3, label="Records after duplicates removed\n(n = 161)",
+  add_node(x = 0, y = 3, label="Records after duplicates removed\n(n = 162)",
            shape="rectangle",
            width=2.5) %>%
-  add_node(x = 0, y = 2, label="Records screened\n(n = 161)",
+  add_node(x = 0, y = 2, label="Records screened\n(n = 162)",
            shape="rectangle",
            width=2) %>%
   add_node(x = 3, y = 2, label="Records excluded\n(n = 115)",
            shape="rectangle",
            width=2) %>%
-  add_node(x = 0, y = 1, label="Full-text articles assessed\nfor eligibility\n(n = 46)",
+  add_node(x = 0, y = 1, label="Full-text articles assessed\nfor eligibility\n(n = 47)",
            shape="rectangle",
            width=2,
            height=0.6) %>%
@@ -39,7 +39,7 @@ create_graph() %>%
   add_edge(4,5) %>%
   add_edge(6,7) %>%
   add_edge(6,8) %>%
-  select_edges() %>%
+  # select_edges() %>%
   add_global_graph_attrs(
     attr = "splines",
     value = "spline",

### 2018-02-23 R/refracrory_graph.R

@@ -65,13 +65,13 @@ g <- create_graph() %>%
   add_edge(8,12) %>%
   add_edge(9,12) %>%
   add_edge(11,12) %>%
-  select_edges() %>%
+  # select_edges() %>%
   add_global_graph_attrs(value="black",
                          attr = "fontcolor",
                          attr_type = "node") %>%
   add_global_graph_attrs(value="ortho",
                          attr = "splines",
                          attr_type = "graph")
-  render_graph(g)
+  # render_graph(g)
   export_graph(g,file_name = "~/Documents/refractoryANA/analysis/figures/algo.png", file_type = "png", title = NULL,
                 width = NULL, height = NULL)

### 2018-02-23 analysis/paper/paper.Rmd
references corrected

@@
 bibliography: references.bib
-csl: erci.csl
+csl: expert-review-of-clinical-immunology.csl
 abstract: |
